---
title: "Neighborhood Composition Analaysis"
---

# Set up

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Saez/workflowr_projects/mibi_analysis/")
```

Loaded packages.

```{r}
suppressPackageStartupMessages(library(mistyR))
suppressPackageStartupMessages(library(future))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(factoextra))
plan("multisession", workers=14)
```

Loading data generated in MIBI_Analysis.Rmd

```{r}
meta <- readRDS("data/meta_data.RDS")
all.expr <- readRDS("data/expression.RDS")
meta.smp <- readRDS("data/sample_meta.RDS")
```

# Introduction

We will use MISTy to try to explain the identify of a given cell using the
identify of the neighboring cells. Therefore, we one-hot encode the cell types,
bypass modeling of the intraview, and only model the paraview consisting of
the 50 nearest-neigbors.

# Generating MISTy Views

Repair the class names and select the relevant columns. For the first 6
rows we get:

TODO: Should we exclude "Other Imunne" class from the analysis?

```{r}
data <- meta %>%
  filter(Group != "Unidentified") %>%
  mutate(Class = case_when(
    Group %in% c("Keratin-positive tumor", "Tumor") ~ "Tumor",
    Group == "Mesenchymal-like" ~ "Mesenchymal",
    Group == "Immune" ~ immuneGroup,
    TRUE ~ Group
  )) %>% select(SampleID, Class, row, col) %>%
  mutate(Class = factor(Class))

data %>% slice_head(n=6)
```

Next up, one-hot encoding of the cell types. Again the first six rows.

```{r}
expr <- mltools::one_hot(data.table::as.data.table(data)) %>%
  select(-c(row, col)) %>%
  rename_all(make.names)
colnames(expr) <- colnames(expr) %>% str_replace_all("_", ".")
expr %>% slice_head(n=6)
```

```{r include=FALSE, eval=FALSE}
# This code chunk is used to show a certain bug.

# Get Expression Values and Postions
raw.expr.smp <- unique(data$SampleID) %>%
  set_names() %>%
  map(function(smp) {
    expr %>%
      filter(SampleID == smp) %>%
      select(-SampleID)
  })
raw.expr.smp$`1` %>%
  slice_head(n=6)

pos.smp <- unique(data$SampleID) %>%
  set_names() %>%
  map(function(smp) {
    data %>%
      filter(SampleID == smp) %>%
      select(c(row, col))
  })
pos.smp$`1` %>%
  slice_head(n=6)

# Compute the views
misty.views.smp <- map2(raw.expr.smp, pos.smp, function(expr, pos) {
  expr %>%
    create_initial_view() %>%
    add_paraview(positions = pos, l = 50, zoi = 0, family = "constant")
})

# Running MISTy
result.folders <- map2(misty.views.smp, 
                      names(misty.views.smp), 
                      function(misty.views, i) {
  print(i)                      
  
  # Check for zero variance in both views and remove if necessary
  vars <- apply(misty.views$intraview$data, MARGIN = 2, FUN = var)
  names.null <- names(vars[which(vars == 0)])
  misty.views$intraview$data <- misty.views$intraview$data %>%
    select(-all_of(names.null))
  
  vars <- apply(misty.views$paraview.50$data, MARGIN = 2, FUN = var)
  names.null <- names(vars[which(vars == 0)])
  print(names.null)
  misty.views$paraview.50$data <- misty.views$paraview.50$data %>%
    select(-all_of(names.null))
  
  misty.views %>%
    run_misty(results.folder = paste0("test/run_", i), bypass.intra = TRUE,
              cv.folds = 10)
})

misty.results <- collect_results(result.folders)

plot_improvement_stats(misty.results, "intra.R2")
plot_improvement_stats(misty.results, "multi.R2")
plot_improvement_stats(misty.results, "gain.R2")

# I get the error in this sample
misty.views <- misty.views.smp[[5]]
views <- misty.views
cv.folds <- 10
method <- "bag"
learner <- "ranger"
n.vars <- NULL
n.learners <- 100
bypass.intra <- TRUE
seed <- 42
cached <- FALSE
ellipsis.args.text <- NULL
... <- NULL
results.folder <- "results"
append <- FALSE
target.subset <- NULL

# I get the error with this class!
target <- "Class_NK"
views <- misty.views

views$intraview$data %>% pull(target) %>% sum

# Problematically if there is only 1 outcome true out of several thousand
# cells, then the coefficient will simply be 0 and the predictions are all 0.
# 
# Now in the folds that happen not to contain that single true outcome
# we basically compare vectors on length n which contain only 0's and 
# caret::R2 return NaN which caused the bug!

n.cells <- 4000
cv.folds <- 10
prediction <- rep(n.cells/cv.folds, n.cells)
outcome <- rep(n.cells/cv.folds, n.cells)
caret::R2(prediction, outcome, formula = "traditional")
```

Select positions.

```{r}
pos.smp <- unique(data$SampleID) %>%
  set_names() %>%
  map(function(smp) {
    data %>%
      filter(SampleID == smp) %>%
      select(c(row, col))
  })
pos.smp$`1` %>%
  slice_head(n=6)
```

Removing all cell types in a given sample if there are fewer than 50 cells of
this type. After that removing all samples that have fewer than 4 cell types 
left. See below which samples are left and how many unique cell types were
retained.

```{r}
expr.smp.raw <- unique(data$SampleID) %>%
  set_names() %>%
  map(function(smp) {
    ret.expr <- expr %>%
      filter(SampleID == smp) %>%
      select(-SampleID)
    
    c.sums <- colSums(ret.expr)
    remove.classes <- names(c.sums)[which(c.sums < 50)]
    
    return(ret.expr %>% select(-all_of(remove.classes)))
  })

# which sample to keep
n.cells <- map_int(expr.smp.raw, ~ ncol(.x))
samples.keep <- names(n.cells[which(n.cells > 3)])
expr.smp <- expr.smp.raw[samples.keep]
pos.smp <- pos.smp[samples.keep]

# last check
map_int(expr.smp, ~ ncol(.x))
```

Computing the MISTy views. In particulat a paraview consisting of the 50
nearest neigbors (`family = "constant"` and `l = 50`)

```{r message=FALSE}
misty.views.smp <- map2(expr.smp, pos.smp, function(expr, pos) {
  expr %>%
    create_initial_view() %>%
    add_paraview(positions = pos, l = 50, zoi = 0, family = "constant")
})
```

# Running MISTy

Run MISTy bypassing the intraview and using 10 CV folds. Per default random
forest is used.

```{r message=FALSE, warning=FALSE}
misty.views.smp$`1`$intraview$data
result.folders <- map2(misty.views.smp, 
                      names(misty.views.smp), 
                      function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("analysis/composition.analysis/smp.", i), 
              bypass.intra = TRUE,
              cv.folds = 10)
})
```

Collectig the results.

```{r message=FALSE}
misty.results <- collect_results(result.folders)
```

# MISTy Results

## Performance

Intra R2:

```{r}
plot_improvement_stats(misty.results, "intra.R2")
```

Multi R2:

```{r}
plot_improvement_stats(misty.results, "multi.R2")
```

Gain R2:

As expected the gain in explained variance equals the multiview explained
variance (since we bypassed modeling of the intraview).

```{r}
plot_improvement_stats(misty.results, "gain.R2")
```

Regarding the error bars let's check for each cell type in how many samples
it was retained after preprocessing. And we see that especially Tregs, dendritic
cells (DC) and natural killer cells (NK) were only retained in frew samples.

TODO: Would it make sense to exlcude those cell types which are present in 
fewer than 5 samples? - See below

```{r}
map(misty.views.smp, ~ colnames(.x$intraview$data)) %>%
  unlist %>%
  table %>%
  data.frame() %>%
  rename("Class" = ".") %>%
  ggplot() +
    geom_bar(aes(x=reorder(Class, -Freq), y=Freq), stat="identity") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Let's look at the gain in R2 accross all samples and targets.

```{r fig.width=8, fig.height=8}
misty.results$improvements %>%
  filter(measure == "gain.R2") %>%
  mutate(sample = str_extract(sample, "(?<=smp.)[0-9]+")) %>%
  select(-measure) %>%
  pivot_wider(names_from = target) %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample") %>%
  as.matrix() %>%
  # removing clustering so we can keep NAs
  pheatmap::pheatmap(cluster_rows=FALSE, cluster_cols=FALSE)
```

What I find interesting is that, in samples 40, 35, 28, 16, 17, and 1, we 
have a pretty high gain in R2 both for B and for tumore cells. What do those 
samples have in common?

Looking at their immages, these samples belong all to the class of rather
compartimentalized TMEs.

```{r fig.width=10, fig.height=6}
meta %>%
  filter(SampleID %in% c(40, 35, 28, 16, 17, 1)) %>%
  ggplot() +
  geom_point(aes(x = row, y = col, col = Group), size = .1) +
  facet_wrap(~ SampleID, ncol=3) + 
  guides(colour = guide_legend(override.aes = list(size = 4)))
```

We can also see this difference comparing the mixing score of those samples
to the rest (excluding cold samples for this comparison).

```{r}
meta.smp %>%
  filter(Classification != "cold") %>%
  mutate(Class = ifelse(SampleID %in% c(40, 35, 28, 16, 17, 1), "In", "Out")) %>%
  ggplot() +
  geom_boxplot(aes(x=Class, y=ratio.tumor.immune)) +
  lims(y= c(0, 150))
```

Let's also have a look at the performance signature returned by MISTy. Since
it actually only contains data for tumor and macrophages we do not even
need to do a PCA.

We can see that the gain in R2 for tumor cells is much higher in comparitmentalized
samples and for macrophages it is somewhat the same except for the fact
that one mixed sample has a much higher gain.

```{r}
perf.sig <- extract_signature(misty.results = misty.results, 
                              type = "performance") %>%
  mutate(SampleID = str_extract(sample, "(?<=smp.)[0-9]+")) %>%
  select(c(SampleID, ends_with("gain.R2")))

ggplot(perf.sig) +
geom_point(aes(x=Class.Macrophages_gain.R2, 
               y=Class.Tumor_gain.R2,
               col=perf.sig %>% left_join(meta.smp, by="SampleID") %>%
                   pull(Classification))) +
  labs(col = "Class")
```

## Contributions

Especially for predicting tumor cell identity the para view is useful, which
fits to the high gain in R2 for tumor cells we have seen before across many
samples.

```{r}
plot_view_contributions(misty.results)
```

## Importances

```{r}
plot_interaction_heatmap(misty.results, "para.50", cutoff = .25)
```

Plotting it using `pheatmap` and adding P or T for predictor and target
to the cell type names. 

```{r}
misty.results$importances %>%
  filter(Importance > 0) %>%
  filter(view == "para.50") %>%
  group_by(Predictor, Target) %>%
  summarise(imp = mean(Importance, na.rm=TRUE, .groups="keep")) %>%
  mutate(Predictor = paste0("P.", Predictor),
         Target = paste0("T.", Target)) %>%
  pivot_wider(names_from = Target, values_from = imp) %>%
  column_to_rownames(var = "Predictor") %>%
  as.matrix() %>% 
  t() %>%
  pheatmap::pheatmap(cluster_rows=FALSE, cluster_cols=FALSE)
```

We can see that dendritic cells are associated with tumor cells. 

# Running MISTy Subset

Next up we want to run MISTy looking only at a subset of the data to
have more meaningful signatures.

We will simply remove Tregs and NK and merge Class_DC.Mono with Class_DC.
Futhermore, we will remove Other.immune because it is rather difficult
to interpret this group.

```{r}
map(misty.views.smp, ~ colnames(.x$intraview$data)) %>%
  unlist %>%
  table %>%
  data.frame() %>%
  rename("Class" = ".") %>%
  ggplot() +
    geom_bar(aes(x=reorder(Class, -Freq), y=Freq), stat="identity") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
to.remove <- c("Class.Tregs", "Class.DC", "Class.NK", "Class.Other.immune")
misty.views.red <- map(misty.views.smp, function(misty.view) {
  if ("Class.DC" %in% colnames(misty.view$intraview$data)) {
    misty.view$intraview$data <- misty.view$intraview$data %>%
      mutate(Class.DC.Mono = Class.DC.Mono + Class.DC)
    
    misty.view$paraview.50$data <- misty.view$paraview.50$data %>%
      mutate(Class.DC.Mono = Class.DC.Mono + Class.DC)
  }
  
  misty.view$intraview$data <- misty.view$intraview$data %>%
    select(-any_of(to.remove))
  
  misty.view$paraview.50$data <- misty.view$paraview.50$data %>%
    select(-any_of(to.remove))
  
  misty.view
})
```

```{r}
map(misty.views.red, ~ colnames(.x$intraview$data)) %>%
  unlist %>%
  table %>%
  data.frame() %>%
  rename("Class" = ".") %>%
  ggplot() +
    geom_bar(aes(x=reorder(Class, -Freq), y=Freq), stat="identity") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Now we can actually run MISTy.

```{r message=FALSE, warning=FALSE}
red.result.folders <- map2(misty.views.red, 
                      names(misty.views.red), 
                      function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("analysis/red.composition.analysis/smp.", i), 
              bypass.intra = TRUE,
              cv.folds = 10)
})
```

```{r message=FALSE}
red.misty.results <- collect_results(red.result.folders)
```

# MISTy Results Subset

```{r}
plot_improvement_stats(red.misty.results, "gain.R2")
```

```{r}
plot_view_contributions(misty.results)
```

```{r}
plot_interaction_heatmap(red.misty.results, "para.50", cutoff = .25)
```


```{r}
# Can be used to make a custom color palette
makeColorRampPalette <- function(colors, cutoff.fraction, num.colors.in.palette)
{
  stopifnot(length(colors) == 5)
  ramp1 <- colorRampPalette(colors[1:2])(num.colors.in.palette * 
                                           cutoff.fraction)
  ramp2 <- colorRampPalette(colors[3:5])(num.colors.in.palette *
                                           (1 - cutoff.fraction))
  return(c(ramp1, ramp2))
}

cols <- makeColorRampPalette(c("white", "white", "blue", "yellow", "red"),
                             cutoff.fraction = .5,
                             num.colors.in.palette = 100)

red.misty.results$importances %>%
  filter(Importance > 0) %>%
  filter(view == "para.50") %>%
  group_by(Predictor, Target) %>%
  summarise(imp = mean(Importance, na.rm=TRUE, .groups="keep")) %>%
  mutate(Predictor = paste0("P.", Predictor),
         Target = paste0("T.", Target)) %>%
  pivot_wider(names_from = Target, values_from = imp) %>%
  column_to_rownames(var = "Predictor") %>%
  as.matrix() %>% 
  pheatmap::pheatmap(cluster_rows=FALSE, cluster_cols=FALSE)
```

# MISTy Without Tumor

Now we could actually run the same analyis excluding tumor cells.

```{r}
to.remove <- c("Class.Tregs", "Class.DC", "Class.NK", "Class.Other.immune",
               "Class.Tumor")
misty.views.notumor <- map(misty.views.smp, function(misty.view) {
  if ("Class.DC" %in% colnames(misty.view$intraview$data)) {
    misty.view$intraview$data <- misty.view$intraview$data %>%
      mutate(Class.DC.Mono = Class.DC.Mono + Class.DC)
    
    misty.view$paraview.50$data <- misty.view$paraview.50$data %>%
      mutate(Class.DC.Mono = Class.DC.Mono + Class.DC)
  }
  
  misty.view$intraview$data <- misty.view$intraview$data %>%
    select(-any_of(to.remove))
  
  misty.view$paraview.50$data <- misty.view$paraview.50$data %>%
    select(-any_of(to.remove))
  
  misty.view
})
```

```{r}
map(misty.views.notumor, ~ colnames(.x$intraview$data)) %>%
  unlist %>%
  table %>%
  data.frame() %>%
  rename("Class" = ".") %>%
  ggplot() +
    geom_bar(aes(x=reorder(Class, -Freq), y=Freq), stat="identity") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r message=FALSE, warning=FALSE}
notumor.result.folders <- map2(misty.views.notumor, 
                      names(misty.views.notumor), 
                      function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = 
                paste0("analysis/notumor.composition.analysis/smp.", i), 
              bypass.intra = TRUE,
              cv.folds = 10)
})
```

```{r message=FALSE}
notumor.misty.results <- collect_results(notumor.result.folders)
```

```{r}
plot_improvement_stats(notumor.misty.results, "gain.R2")
```

Results stratified by classification.

Based on these results, I will only look at the targets below for which
we got a gain in R2, namely: B, CD4, CD8, DC.Mono, Endothelial,
Macrophages, Mesenchymal, Neutrophils

```{r}
notumor.misty.results$improvements %>%
  mutate(sample = str_extract(sample, "(?<=smp.)[0-9]+")) %>%
  filter(measure == "gain.R2") %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  ggplot() +
  geom_boxplot(aes(x=target, y=value, col=Classification)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```

And we see that we can predict the identify of cells much better in
compartimentalized tumors.

```{r}
plot_view_contributions(notumor.misty.results)
```

```{r}
plot_interaction_heatmap(notumor.misty.results, "para.50", cutoff = .4)
```


```{r}
notumor.misty.results$importances %>%
  filter(Importance > 0) %>%
  filter(view == "para.50") %>%
  group_by(Predictor, Target) %>%
  summarise(imp = mean(Importance, na.rm=TRUE, .groups="keep")) %>%
  mutate(Predictor = paste0("P.", Predictor),
         Target = paste0("T.", Target)) %>%
  filter(!(Target %in% c("T.Class_CD3.T", "T.Class_Mono.Neu"))) %>%
  pivot_wider(names_from = Target, values_from = imp) %>%
  column_to_rownames(var = "Predictor") %>%
  as.matrix() %>% 
  pheatmap::pheatmap(cluster_rows=FALSE, cluster_cols=FALSE)
```

Only mixed samples:

```{r}
notumor.misty.results$importances %>%
  mutate(sample = str_extract(sample, "(?<=smp.)[0-9]+")) %>%
  filter(Importance > 0) %>%
  filter(view == "para.50") %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  filter(Classification == "mixed") %>%
  select(sample, Predictor, Target, Importance) %>%
  group_by(Predictor, Target) %>%
  summarise(imp = mean(Importance, na.rm=TRUE, .groups="keep")) %>%
  mutate(Predictor = paste0("P.", Predictor),
         Target = paste0("T.", Target)) %>%
  filter(!(Target %in% c("T.Class_CD3.T", "T.Class_Mono.Neu"))) %>%
  pivot_wider(names_from = Target, values_from = imp) %>%
  column_to_rownames(var = "Predictor") %>%
  as.matrix() %>% 
  pheatmap::pheatmap(cluster_rows=FALSE, cluster_cols=FALSE)
```

Only compartimentalized samples.

```{r}
notumor.misty.results$importances %>%
  mutate(sample = str_extract(sample, "(?<=smp.)[0-9]+")) %>%
  filter(Importance > 0) %>%
  filter(view == "para.50") %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  filter(Classification == "compartimentalized") %>%
  select(sample, Predictor, Target, Importance) %>%
  group_by(Predictor, Target) %>%
  summarise(imp = mean(Importance, na.rm=TRUE, .groups="keep")) %>%
  mutate(Predictor = paste0("P.", Predictor),
         Target = paste0("T.", Target)) %>%
  filter(!(Target %in% c("T.Class_CD3.T", "T.Class_Mono.Neu"))) %>%
  pivot_wider(names_from = Target, values_from = imp) %>%
  column_to_rownames(var = "Predictor") %>%
  as.matrix() %>% 
  pheatmap::pheatmap(cluster_rows=FALSE, cluster_cols=FALSE)
```

Looking at the actual sample pictures whether these importances make sense.

First, B and CD4 T cells in mixed TMEs

```{r fig.width=10, fig.height=8}
meta %>%
  mutate(SampleID = as.character(SampleID)) %>%
  left_join(meta.smp) %>%
  filter(Classification == "mixed") %>%
  mutate(class = case_when(
    Group %in% c("Keratin-positive tumor", "Tumor") ~ "Tumor",
    !(immuneGroup %in% c("B", "CD4 T")) ~ "Ignore",
    TRUE ~ immuneGroup
  )) %>%
  filter(class != "Ignore") %>%
  ggplot() +
  geom_point(aes(x=row, y=col, col=class), size=.1) +
  facet_wrap(~ SampleID) +
  guides(colour = guide_legend(override.aes = list(size = 2)))
```

Second, B and CD4 T cells in compartimentalized TMEs

```{r fig.width=10, fig.height=8}
meta %>%
  mutate(SampleID = as.character(SampleID)) %>%
  left_join(meta.smp) %>%
  filter(Classification == "compartimentalized") %>%
  mutate(class = case_when(
    Group %in% c("Keratin-positive tumor", "Tumor") ~ "Tumor",
    !(immuneGroup %in% c("B", "CD4 T")) ~ "Ignore",
    TRUE ~ immuneGroup
  )) %>%
  filter(class != "Ignore") %>%
  ggplot() +
  geom_point(aes(x=row, y=col, col=class), size=.1) +
  facet_wrap(~ SampleID) +
  guides(colour = guide_legend(override.aes = list(size = 2)))
```

And in fact one can see that CD4 T and B cells colocalize.


