---
title: "Exploratory Data Analysis MiBi"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

# Set up

Messages will be hidden and the output collapsed to make the Rmd more clear.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, message = FALSE)
```

Loaded Packages.

```{r}
library(mistyR)
library(future)
plan("multisession", workers=14)
library(tidyverse)
```

# Loading and Processing Data

```{r}
input <- read_csv("data/cellData.csv")

# See description in Readme
group.trans <- c("1" = "Unidentified", "2" = "Immune", "3" = "Endothelial", 
                 "4" = "Mesenchymal-like", "5" = "Tumor", 
                 "6" = "Keratin-positive tumor")

immune.group.trans <- c("0" = "Non-Immune?", "1" = "Tregs", "2" = "CD4 T", "3" = "CD8 T", 
                        "4" = "CD3 T", "5" = "NK", "6" = "B",
                        "7" = "Neutrophils", "8" = "Macrophages", 
                        "9" = "DC", "10" = "DC/Mono", "11" = "Mono/Neu", 
                        "12" = "Other immune")

raw_data <- input %>%
  mutate(Group = group.trans[as.character(Group)]) %>%
  mutate(immuneGroup = immune.group.trans[as.character(immuneGroup)])
  
raw_data %>% slice_head(n=6)
```

Add coordinates of the cells.

There are different ways to get the center of mass. One of the is implemented by
scipy, see here: https://docs.scipy.org/doc/scipy/reference/generated/scipy.ndimage.center_of_mass.html#scipy.ndimage.center_of_mass.

```{r}
# Coordinates extracted using scipy ndimage.measurements.center_of_mass
coordinates_raw <- read_csv("data/coordinates.csv")
coordinates <- coordinates_raw %>%
  # changed row and col here, according to graphics later
  mutate(row = as.numeric(str_extract(coordinate, "[.\\d]+(?=,)")),
         col = as.numeric(str_extract(coordinate, "[.\\d]+(?=\\))"))
         ) %>%
  select(-coordinate) %>%
  rename(cell_id = ...1, sample_id = id) %>%
  mutate(cell_id = cell_id + 1) # control for python vs. R, first index
coordinates %>% slice_head(n=6)
```

Add coordinates to the meta_data

```{r}
data <- raw_data %>%
  inner_join(coordinates, by = c("cellLabelInImage" = "cell_id", 
                                "SampleID" = "sample_id")) %>%
  drop_na() # put drop NA here (so it works at least)

# Check for NAs
stopifnot(sum(map_int(colnames(data), ~ sum(is.na(data[[.x]])))) == 0)

meta <- data %>%
  select(c(1:3, 53:59))
meta %>% slice_head(n=6)

expr <- data %>%
  select(4:52)
expr %>% slice_head(n=6)
```

Alternatively we can do it in R as shown below. However, this takes
much longer than simply running the code in Python (about 100X difference?).

```{r}
if ("alternative_coordinates.RDS" %in% list.files("data")) {
  alternative.coord <- readRDS("data/alternative_coordinates.RDS")
} else {
  # maybe make the matrix sparse to speed things up?
  library(tiff)
  lbs <- c(1:29, 31:41)
  alternative.coord <- furrr::future_map_dfr(lbs, function(id) {
    print(id)
    # Read in raw matrix (in tiff format)
    tiff <- readTIFF(paste0("/home/philipp/data/saez/p", id, "_labeledcellData.tiff"),
                     as.is = TRUE)
    seq.rows <- seq_len(nrow(tiff))
    seq.cols <- seq_len(ncol(tiff))
    ncells <- cells.sample.dict[id]
    map_dfr(seq_len(ncells), function(i) {
      if (i %% 100 == 0) print(i)
      # Convert to binary matrix with TRUE and FALSE
      binary <- (tiff == i)
      s <- sum(binary)
      # Calculate center of mass
      c(id = id, 
        i = i,
        x.center = sum(seq.rows * rowSums(binary)) / s,
        y.center = sum(seq.cols * colSums(binary)) / s
      )
    })
  })
  saveRDS(alternative.coord, file = "data/alternative_coordinates.RDS")
}
```

# Metadata

Number of cells per sample.

```{r}
meta %>%
  count(SampleID) %>%
  ggplot() +
  geom_bar(aes(x=SampleID, y=n), stat = "identity")

cells.sample <- input %>% count(SampleID)
cells.sample.dict <- cells.sample$n
names(cells.sample.dict) <- cells.sample$SampleID
cells.sample.dict
```

Number of "cell types" over all samples.

```{r}
meta %>%
  count(Group) %>%
  arrange(desc(n))
```

Number of "immune cell types" over all samples.

```{r}
meta %>%
  count(immuneGroup) %>%
  arrange(desc(n))
```

# Distribution of the markers

Histograms per marker over all samples.

```{r fig.height=12, fig.width=18}
expr %>%
  pivot_longer(cols = everything(), names_to = "marker", values_to = "expression") %>%
  ggplot() +
  geom_histogram(aes(x=expression), bins = 100) +
  facet_wrap(~ marker, scales = "free")
```

```{r fig.height=12, fig.width=18}
expr %>%
  pivot_longer(cols = everything(), names_to = "marker", values_to = "expression") %>%
  ggplot() +
  geom_histogram(aes(x=expression), bins = 100) +
  facet_wrap(~ marker) +
  scale_x_log10()
```

Looking at the boxplot fo the expression per marker over all samples, ignoring
outliers which are defined as points which includes all points outside of 
median ± 1.5 * IQR (interquantile range)

```{r}
expr %>%
  pivot_longer(cols = everything(), names_to = "marker", values_to = "expression") %>%
  ggplot() +
  geom_boxplot(aes(y=expression, x=marker), outlier.shape = NA) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  coord_cartesian(ylim = c(-5, 5))
```

# Visualizing the cell types in the coordinates

```{r fig.height=12, fig.width=18}
meta %>%
  ggplot() +
  geom_point(aes(x = row, y = col, col = Group), size=.1) +
  facet_wrap(~ SampleID)
```

Let's visualize sample 40 for example in one slide.

```{r fig.height=6, fig.width=8}
meta %>%
  filter(SampleID == 40) %>%
  ggplot() +
  geom_point(aes(x = row, y = col, col = Group), size=.4)
```

And also visualize the tiff for that slide (or sample).

```{r fig.height=6, fig.width=8}
library(tiff)
id = 40
tiff <- readTIFF(paste0("data/raw_tiffs/p", 
                        id, "_labeledcellData.tiff"), as.is = TRUE)
binary <- apply(tiff, c(1,2), function(n) !(n %in% c(0,1)))
binary.df <- as.data.frame(binary)
rownames(binary.df) <- seq_len(nrow(binary))
colnames(binary.df) <- seq_len(ncol(binary))

df <- binary.df %>% 
  rownames_to_column(var = "row") %>%
  pivot_longer(cols = !(row), names_to = "col") %>%
  mutate(row = as.numeric(row), col = as.numeric(col))

df %>%
  ggplot() +
  geom_point(aes(x=row, y=col, col=value), size=.1) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank())
```

Visualize cells and tiff together.

```{r fig.height=8, fig.width=10}
s.40 <- meta %>%
  filter(SampleID == 40)

tiff.40 <- df %>%
  filter(value) %>%
  slice_sample(., n = nrow(.)/10)

ggplot() +
  geom_point(data = tiff.40, aes(x=row, y=col), col="black", 
             size = 0.1, stroke = 0, shape = 16) +
  geom_point(data = s.40, aes(x=row, y=col, col=Group), size = 1) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank())
```

As Jovan suggested I will look at a subset of the data. And we can see
that the center of mass (colored point) mostly matches the cell boundaries,
but not always!

```{r fig.height=8, fig.width=10}
c.max <- dim(binary)[1]
top_left <- list(row = 0, col = c.max)
bottom_right <- list(row = 600, col = c.max-600)

s.40 <- meta %>%
  filter(SampleID == 40) %>%
  filter(row >= top_left$row, row <= bottom_right$row,
         col >= bottom_right$col, col <= top_left$col)

tiff.40 <- df %>%
  filter(value) %>%
  filter(row >= top_left$row, row <= bottom_right$row,
         col >= bottom_right$col, col <= top_left$col)

ggplot() +
  geom_point(data = tiff.40, aes(x=row, y=col), col="black", 
             size = 0.1, stroke = 0, shape = 16) +
  geom_point(data = s.40, aes(x=row, y=col, col=Group), size = 3) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank())
```

# Run MISTy on all samples

Preparation

```{r}
expr.smp <- unique(meta$SampleID) %>%
  set_names(paste0("s", .)) %>%
  map(function(id) {
    ret.expr <- expr %>%
      filter(meta$SampleID == id) %>%
      # Select only proteins
      select(11:47)
      
    colnames(ret.expr) <- make.names(colnames(ret.expr))
    ret.expr
  })

sum(map_int(names(expr.smp$s40), ~ sum(is.na(expr.smp$s40[.x]))))

coord.smp <- unique(meta$SampleID) %>%
  set_names(paste0("s", .)) %>%
  map(function(id) {
    meta %>%
      filter(meta$SampleID == id) %>% 
      select(row, col)
  })

sum(map_int(names(coord.smp$s40), ~ sum(is.na(coord.smp$s40[.x]))))

nrow(expr.smp$s40) == nrow(coord.smp$s40)
```

Generating the misty.views

```{r}
if ("misty_views_smp.RDS" %in% list.files("data")) {
  misty.views.smp <- readRDS("data/misty_views_smp.RDS")
} else {
  cv.folds = 10
  misty.views.smp <- map2(expr.smp, coord.smp, function(expr, coord) {
    
    # Check for zero variance (otherwise MISTy throws an error)
    target.var <- apply(expr, 2, stats::sd, na.rm = TRUE)
    expr <- expr %>% select(-names(which(target.var == 0)))
    
    # Check for how many uniqzue values
    target.unique <- colnames(expr) %>%
      purrr::set_names() %>%
      purrr::map_int(~ length(unique(expr %>% pull(.x))))
    expr <- expr %>% select(names(target.unique[target.unique > cv.folds]))
    
    # Create views and run MISTy
    create_initial_view(expr) %>%
      add_juxtaview(positions = coord) %>%
      add_paraview(positions = coord, l = 150, zoi = 15)
  })
  names(misty.views.smp) <- names(expr.smp)
  saveRDS(misty.views.smp, "data/misty_views_smp.RDS")
}
```


```{r}
if ("misty_results_smp.RDS" %in% list.files("data")) {
  results.smp <- readRDS("data/misty_results_smp.RDS")
} else {
  cv.folds = 10
  results.smp <- misty.views.smp %>%
    set_names() %>%
    map(function(smp) {
      smp %>%
        run_misty(cv.folds = cv.folds) %>%
        collect_results()
    })
  names(results.smp) <- names(expr.smp)
  saveRDS(results.smp, "data/misty_results_smp.RDS")
}
```

```{r}
source("code/custom_plotting.R")
```

<details>
<summary>Click to view all diagnostic plots</summary>

```{r fig.height=14, fig.width=20}
walk2(results.smp, names(results.smp), function(res, name) {
  plot_diagnostics(res, name, low=.5)
})
```

```{r}
plot_improvement_stats(results.smp$s24)
print(plot_interaction_heatmap(results.smp$s24, "para.150"))
```

</details>

```{r}
if ("misty_results_earth_smp.RDS" %in% list.files("data")) {
  earth.results.smp <- readRDS("data/misty_results_earth_smp.RDS")
} else {
  cv.folds = 10
  earth.results.smp <- misty.views.smp %>%
    set_names() %>%
    map(function(smp) {
      smp %>%
        run_misty(cv.folds = cv.folds, 
                  method = "bag", 
                  learner = "earth",
                  n.learners = 100) %>%
        collect_results()
    })
  names(earth.results.smp) <- names(expr.smp)
  saveRDS(earth.results.smp, "data/misty_results_earth_smp.RDS")
}
```

```{r fig.height=14, fig.width=20}
plot_diagnostics(earth.results.smp$s1, name="s1")
```

<details>
<summary>Click to view all diagnostic plots</summary>

```{r fig.height=14, fig.width=20}
walk2(earth.results.smp, names(earth.results.smp), function(res, name) {
  plot_diagnostics(res, name, low=.5)
})
```

</details>

# Aggregating results

```{r}
targets <- unique(results.smp$s1$improvements$target)

ranger.measures <- map_dfr(targets, function(tar) {
  map2_dfr(results.smp, names(results.smp), function(smp, n) {
    smp$improvements %>%
      filter(target == tar) %>%
      filter(measure %in% c("intra.R2", "multi.R2", "gain.R2")) %>%
      mutate(sample = n)
  })
})
```

```{r}
ranger.measures %>%
  filter(measure == "gain.R2") %>%
  group_by(target) %>%
  summarise(mean.gain = mean(value)) %>%
  ggplot() +
  geom_bar(aes(x = target, y = mean.gain), stat="identity") +
  theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1)) 

ranger.measures %>%
  filter(measure == "gain.R2") %>%
  ggplot() +
  geom_boxplot(aes(x = target, y = value)) +
  theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1)) 
```

```{r}
earth.measures <- map_dfr(targets, function(tar) {
  map2_dfr(earth.results.smp, names(earth.results.smp), function(smp, n) {
    smp$improvements %>%
      filter(target == tar) %>%
      filter(measure %in% c("intra.R2", "multi.R2", "gain.R2")) %>%
      mutate(sample = n)
  })
})
```

```{r}
earth.measures %>%
  filter(measure == "gain.R2") %>%
  group_by(target) %>%
  summarise(mean.gain = mean(value)) %>%
  ggplot() +
  geom_bar(aes(x = target, y = mean.gain), stat="identity") +
  theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1)) 

earth.measures %>%
  filter(measure == "gain.R2") %>%
  ggplot() +
  geom_boxplot(aes(x = target, y = value)) +
  theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1)) 
```

Comparison on ranger and earth results.

```{r fig.width=8, fig.height=8}
tmp1 <- ranger.measures %>% mutate(method = "ranger")
tmp2 <- earth.measures %>% mutate(method = "earth")
all <- rbind(tmp1, tmp2)
all %>%
  filter(measure == "multi.R2") %>%
  ggplot() +
  geom_boxplot(aes(x=target, y=value)) +
  facet_wrap(~ method, nrow=2) +
  theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1)) 
```

```{r fig.width=10, fig.height=6}
m <- "gain.R2"
all %>%
  mutate(new_target = paste0(target, "_", method)) %>%
  filter(measure == m) %>%
  ggplot() +
  geom_boxplot(aes(x=new_target, y=value, col=method)) +
  labs(title = m) +
  theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5))

all %>%
  mutate(new_target = paste0(target, "_", method)) %>%
  filter(measure == m) %>%
  ggplot() +
  geom_boxplot(aes(x=new_target, y=value, col=method)) +
  labs(title = m) +
  theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5)) +
  lims(y = c(-2.5,2.5))

m <- "multi.R2"
all %>%
  mutate(new_target = paste0(target, "_", method)) %>%
  filter(measure == m) %>%
  ggplot() +
  geom_boxplot(aes(x=new_target, y=value, col=method)) +
  labs(title = m) +
  theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5))

m <- "intra.R2"
all %>%
  mutate(new_target = paste0(target, "_", method)) %>%
  filter(measure == m) %>%
  ggplot() +
  geom_boxplot(aes(x=new_target, y=value, col=method)) +
  labs(title = m) +
  theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5))
```



