---
title: "Exploratory Data Analysis MIBI with Helper Function"
---

# Set up

Messages will be hidden and the output collapsed to make the Rmd more clear.

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Saez/mibi_analysis/")
```

Loaded Packages.

```{r}
suppressPackageStartupMessages(library(mistyR))
suppressPackageStartupMessages(library(future))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(factoextra))
suppressPackageStartupMessages(library(tiff))
plan("multisession", workers=14)
```

# Loading and Processing Data

The columns Group and ImmuneGroup in the cell metadata are translated from
numbers to their actual meaning.

```{r}
input <- read_csv("data/cellData.csv", show_col_types=FALSE)

# See description in Readme
group.trans <- c("1" = "Unidentified", "2" = "Immune", "3" = "Endothelial", 
                 "4" = "Mesenchymal-like", "5" = "Tumor", 
                 "6" = "Keratin-positive tumor")

immune.group.trans <- c("0" = "Non-Immune?", "1" = "Tregs", "2" = "CD4 T", 
                        "3" = "CD8 T", "4" = "CD3 T", "5" = "NK", "6" = "B",
                        "7" = "Neutrophils", "8" = "Macrophages", 
                        "9" = "DC", "10" = "DC/Mono", "11" = "Mono/Neu", 
                        "12" = "Other immune")

raw_data <- input %>%
  mutate(Group = group.trans[as.character(Group)]) %>%
  mutate(immuneGroup = immune.group.trans[as.character(immuneGroup)])
  
raw_data %>% slice_head(n=6)
```

Next we add the coordinates of the cells.

```{r}
lbs <- c(1:29, 31:41) # actual samples
coord <- furrr::future_map_dfr(lbs, function(id) {
  # Read in raw matrix (in tiff format)
  tiff <- readTIFF(paste0("/home/philipp/data/saez/mibi/p", id, "_labeledcellData.tiff"),
                   as.is = TRUE)
  seq.rows <- seq_len(nrow(tiff))
  seq.cols <- seq_len(ncol(tiff))
  # important: map over all unique values here! (but I removed 1 and 2)
  cell.ids <- unique(as.vector(tiff))[-which(unique(as.vector(tiff)) %in% c(0,1))]
  map_dfr(cell.ids, function(i) {
    # Convert to binary matrix with TRUE and FALSE
    binary <- (tiff == i)
    s <- sum(binary)
    # Calculate center of mass
    c(id = id, 
      i = i,
      x.center = sum(seq.rows * rowSums(binary)) / s,
      y.center = sum(seq.cols * colSums(binary)) / s
    )
  })
})
```

Then, join the coordinates with the meta data.

```{r}
data <- raw_data %>%
  inner_join(coord, by = c("SampleID" = "id", "cellLabelInImage" = "i")) %>%
  rename(row = x.center, col = y.center)
```

Split the data into expression matrix and meta data.

```{r}
meta <- data %>%
  select(c(1:3, 53:59))
meta %>% slice_head(n=6)

expr <- data %>%
  select(4:52)
expr %>% slice_head(n=6)
```

Patient data provided with the raw data.

```{r}
patient.info <- read_csv("data/patient_class.csv",
                         col_names = c("SampleID", "Classification"),
                         show_col_types=FALSE)
patient.info %>% slice_head(n=6)

patient.info.paper <- readxl::read_xlsx("data/mmc2.xlsx", 
                                        skip = 1, col_names = TRUE)
patient.info.paper %>% slice_head(n=6)
```

Merging the meta data with other measures.

```{r}
# Generating Sample Meta Data
immune.counts <- meta %>%
  group_by(SampleID) %>%
  count(Group) %>%
  pivot_wider(names_from = Group, values_from = n,
              names_prefix = "Count.")

total.counts <- meta %>%
  count(SampleID) %>%
  rename(Count.Total = n)

sample.meta <- patient.info %>%
  filter(SampleID != 30) %>%
  inner_join(immune.counts, by = "SampleID") %>%
  inner_join(total.counts, by = "SampleID") %>%
  # see paper Figure 3, Panel C
  mutate(Classification = 
         case_when(Classification == 0 ~ "mixed",
                   Classification == 1 ~ "compartimentalized",
                   Classification == 2 ~ "cold")) %>% 
  mutate(SampleID = as.character(SampleID)) %>%
  left_join(patient.info.paper, by = c("SampleID" = "InternalId"))

sample.meta %>% slice_head(n=6)
```

# Running MISTy

## Preparation

Select only proteins from the expression matrix and check for NAs.

Also we need to check the variance for each marker in each same and
exclude the marker if the variance is 0. The same is true for markers
whose number of unique values is lower than the desired number of cv
folds (here 10).

```{r collapse=TRUE}
raw.expr <- unique(meta$SampleID) %>%
  set_names(paste0("s", .)) %>%
  map(function(id) {
    ret.expr <- expr %>%
      filter(meta$SampleID == id) %>%
      select(11:47) }) # Select only proteins

expr.smp <- mistyR.helpers::prepare_markers(raw.expr)

coord.smp <- unique(meta$SampleID) %>%
  set_names(paste0("s", .)) %>%
  map(function(id) {
    meta %>%
      filter(meta$SampleID == id) %>% 
      select(row, col)
  })
```

First we will look at the distribution of the  mean distances to the direct 
neighbors (as determined by the delauney triangulation) to get a feeling for 
which cutoff value should be used for the generation of the juxtaview. The
corresponding distribution is shown below.
(in blue the cutoff - 40 pixels (15.6 microns) - which was used in the end).

```{r}
mean.dists.direct.neighbors <- mistyR.helpers::distances_juxtaview(coord.smp)

mistyR.helpers::plot_distances_juxtaview(mean.dists.direct.neighbors, 100) +
  geom_vline(xintercept = 40, col = "red", alpha = 0.5)
```

However, a more detailed look across the sample (boxplot) shows that the
cutoff may be too low for certain samples. It may be good to increase this
threshold to 50.

```{r}
# boxplot without outliers
mean.dists.direct.neighbors %>%
  mutate(sample = factor(sample, levels = unique(sample))) %>%
  ggplot() +
  geom_boxplot(aes(x=sample, y=mean.distance), outlier.shape = NA) +
  geom_hline(yintercept = 40, col="blue") +
  lims(y = c(0, 100)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5))
```

Furthermore we will look at the distribution of the mean distances to the
10 -, 20 -, 40 -, and 80 closest cells to get a feeling
for which cutoff will be used to generate the paraview 
(in blue the cutoff - 120 pixels (46.8 microns) - which was used in the end).

```{r}
neighbor.dists <- mistyR.helpers::distances_paraview(coord.smp)

mistyR.helpers::plot_distances_paraview(neighbor.dists, 200) +
  geom_vline(xintercept = 120, col = "red", alpha = 0.5)
```

## Generating the MISTy views

Based on the data above, the following parameters (thresholds) were used 
to run MISTy: 
 * Juxtaview: threshold 40,
 * Paraview: theshold 120, zone of indifference: 40

```{r message=FALSE}
misty.views.smp <- map2(expr.smp, coord.smp, function(expr, coord) {
    # Create views and run MISTy
    create_initial_view(expr) %>%
      add_juxtaview(positions = coord, neighbor.thr = 40) %>%
      add_paraview(positions = coord, l = 120, zoi = 40)
  })
names(misty.views.smp) <- names(expr.smp)
```

## Random Forest Implementation

Running with random forest as ML algorithm to build the view-specific models. 

```{r message=FALSE}
cv.folds = 10
ranger.results.folders <- map2(
  misty.views.smp, names(misty.views.smp), function(smp, name) {
    
    smp %>% run_misty(results.folder = paste0("analysis/results/ranger_", name),
                      cv.folds = cv.folds)
  })
ranger.results <- collect_results(ranger.results.folders)
```

# MISTy Results

```{r}
plot_improvement_stats(ranger.results, "gain.R2")
```

```{r}
plot_view_contributions(ranger.results)
```

```{r}
plot_interaction_heatmap(ranger.results, "para.120", clean = TRUE)
```
