---
title: "Exploratory Data Analysis MIBI"
---

# Set up

Messages will be hidden and the output collapsed to make the Rmd more clear.

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Saez/workflowr_projects/mibi_analysis/")
```

Loaded Packages.

```{r}
suppressPackageStartupMessages(library(mistyR))
suppressPackageStartupMessages(library(future))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(factoextra))
plan("multisession", workers=14)
```

# Loading and Processing Data

The columns Group and ImmuneGroup in the cell metadata are translated from
numbers to their actual meaning.

```{r}
input <- read_csv("data/cellData.csv", show_col_types=FALSE)

# See description in Readme
group.trans <- c("1" = "Unidentified", "2" = "Immune", "3" = "Endothelial", 
                 "4" = "Mesenchymal-like", "5" = "Tumor", 
                 "6" = "Keratin-positive tumor")

immune.group.trans <- c("0" = "Non-Immune?", "1" = "Tregs", "2" = "CD4 T", "3" = "CD8 T", 
                        "4" = "CD3 T", "5" = "NK", "6" = "B",
                        "7" = "Neutrophils", "8" = "Macrophages", 
                        "9" = "DC", "10" = "DC/Mono", "11" = "Mono/Neu", 
                        "12" = "Other immune")

raw_data <- input %>%
  mutate(Group = group.trans[as.character(Group)]) %>%
  mutate(immuneGroup = immune.group.trans[as.character(immuneGroup)])
  
raw_data %>% slice_head(n=6)
```

Next we add the coordinates of the cells.

Doing it in R takes quite some time but offers more control than the 
[center of mass function in scipy](https://docs.scipy.org/doc/scipy/reference/generated/scipy.ndimage.center_of_mass.html).

```{r}
if ("coord.RDS" %in% list.files("data")) {
  coord <- readRDS("data/coord.RDS")
} else {
  library(tiff)
  lbs <- c(1:29, 31:41)
  coord <- furrr::future_map_dfr(lbs, function(id) {
    print(id)
    # Read in raw matrix (in tiff format)
    tiff <- readTIFF(paste0("/home/philipp/data/saez/p", id, "_labeledcellData.tiff"),
                     as.is = TRUE)
    seq.rows <- seq_len(nrow(tiff))
    seq.cols <- seq_len(ncol(tiff))
    # important: map over all unique values here! (but I removed 1 and 2)
    cell.ids <- unique(as.vector(tiff))[-which(unique(as.vector(tiff)) %in% c(0,1))]
    map_dfr(cell.ids, function(i) {
      if (i %% 100 == 0) print(i)
      # Convert to binary matrix with TRUE and FALSE
      binary <- (tiff == i)
      s <- sum(binary)
      # Calculate center of mass
      c(id = id, 
        i = i,
        x.center = sum(seq.rows * rowSums(binary)) / s,
        y.center = sum(seq.cols * colSums(binary)) / s
      )
    })
  })
  saveRDS(coord, file = "data/coord.RDS")
}
```

Then, join the coordinates with the meta data.

```{r}
data <- raw_data %>%
  inner_join(coord, by = c("SampleID" = "id", 
                                       "cellLabelInImage" = "i")) %>%
  rename(row = x.center, col = y.center)
```

However, we have some coordiantes for cell IDs which are not in the meta data.
For sample 40 there are for example 572 of those cell IDs which can be 
found in the tiff but not in the meta data. Maybe those spots turned out not 
be cells in the end (segmentation errors?). Or these are the cells with low 
expression across all markers (sum < 0.1).

```{r}
dplyr::setdiff(
  coord %>% filter(id == 40) %>% pull(i),
  data %>% filter(SampleID == 40) %>% pull(cellLabelInImage)
  ) %>%
  length()
```

```{r echo=FALSE}
# Before proceeding with the analysis, check whether we have any NAs.
stopifnot(sum(map_int(colnames(data), ~ sum(is.na(data[[.x]])))) == 0)
```

Split the data into expression matrix and meta data.

```{r}
meta <- data %>%
  select(c(1:3, 53:59))
meta %>% slice_head(n=6)

expr <- data %>%
  select(4:52)
expr %>% slice_head(n=6)
```

# Meta Data

## Cell Counts

Number of cells per sample. Curiously sample 30 is missing.

```{r}
meta %>%
  count(SampleID) %>%
  ggplot() +
  geom_bar(aes(x=SampleID, y=n), stat = "identity")
```

Number of "cell types" over all samples as annotated by the authors or the
study. We mostly have tumor and immune cells. Keratin serves as a marker
for cancer cells.

```{r}
meta %>%
  count(Group) %>%
  arrange(desc(n))
```

Next we will have a closer look at the immune populations. And according
to the number of cells present, macrophages and T-cells are the major
players, followed by B-cells.

```{r}
meta %>%
  count(immuneGroup) %>%
  arrange(desc(n))
```

## Patient/Sample Metadata

Patient data provided with the raw data.

```{r}
patient.info <- read_csv("data/patient_class.csv",
                         col_names = c("SampleID", "Classification"),
                         show_col_types=FALSE)
patient.info %>% slice_head(n=6)
```

```{r}
patient.info.paper <- readxl::read_xlsx("data/mmc2.xlsx", 
                                        skip = 1, col_names = TRUE)

patient.info.paper %>% slice_head(n=6)
```

## Cell Annotation

How were the cells actually annotated? In the paper they used hierachical
clustering (FlowSOM). Here we will only look at the first two principal 
components.

When looking at the first two principal components, we see quite a clear
distinction between tumor and immune cells.
(Random subsetting used to reduce overplotting).

```{r fig.height=4, fig.width=12}
pca <- stats::prcomp(expr)
subs <- sample(nrow(pca$x), .05*nrow(pca$x))
p1 <- ggplot() +
  geom_point(aes(x=seq_len(length(pca$sdev)), y = pca$sdev^2))
p2 <- ggplot() +
  geom_point(aes(x=pca$x[subs,1], pca$x[subs,2], 
                 color = meta$Group[subs]), size = 0.1) +
  guides(colour = guide_legend(override.aes = list(size = 2)))
gridExtra::grid.arrange(p1, p2, ncol=2)
```

Also for the sake of it the will cluster the cells using the louvain
algorithm (graph-based). See the heatmap of the marker distribution accross
the clusters below.

```{r fig.width=6, fig.height=6}
if ("graph.clusters.RDS" %in% list.files("data")) {
  graph.clusters <- readRDS("data/graph.clusters.RDS")
} else {
  # shared nearest neighbor graph
  snn <- bluster::makeSNNGraph(pca$x[,1:20], k = 20)
  # louvain clustering
  graph.clusters <- igraph::cluster_louvain(snn)$membership
  saveRDS(graph.clusters, "data/graph.clusters.RDS")
}

# Making Psuedobulks per cluster
expr.matrix <- as.matrix(
  expr %>%
    select(- c("Background", "P", "Si", "Au", "Na", "Ca", "Ta", "dsDNA",
               "Fe"))
)
pseudobulk <- as.matrix(
  map_dfr(unique(graph.clusters), function(i) {
    colSums(expr.matrix[graph.clusters == i,])
  })
)
rownames(pseudobulk) <- unique(graph.clusters)

# Plotting Heatmap
pheatmap::pheatmap(t(pseudobulk))
```

# Count Data

Histograms per marker over all samples. We can see that that there is a huge
variability in the expression of different markers. The measured entities 
other than proteins (Au, Background, Beta catentin, C, Ca, Fe, Si, Ta) were
later excluded from the analysis with MISTy.

How were the "counts" normalized?

According to the readme.rtf the markers were normalized by dividing the 
counts by the cell size and transforming with arcsinh (inverse hyperbolic sine).
Furthermore in the paper it is stated that after arcsinh transformation the
markers were standard normalized.

```{r fig.height=12, fig.width=18}
expr %>%
  pivot_longer(cols = everything(), names_to = "marker", values_to = "expression") %>%
  ggplot() +
  geom_histogram(aes(x=expression), bins = 100) +
  facet_wrap(~ marker, scales = "free")
```

Looking at the boxplot for the expression per marker over all samples, ignoring
outliers which are defined as points which includes all points outside of 
median ± 1.5 * IQR (interquantile range), we can again see the broad
variability.

```{r}
expr %>%
  pivot_longer(cols = everything(), names_to = "marker", values_to = "expression") %>%
  ggplot() +
  geom_boxplot(aes(y=expression, x=marker), outlier.shape = NA) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  coord_cartesian(ylim = c(-5, 5))
```

Focusing on the measured proteins, the most highly expressed species 
(according to the mean) are:

```{r}
tibble(marker = colnames(expr), mean.expression = colMeans(expr)) %>%
  filter(!(marker %in% c("Ta", "Au", "P", "Si", "C", "Na", "Ca", "Fe", 
                         "dsDNA", "Backgrond"))) %>%
  slice_max(mean.expression, n=6)
```

# Spatial Data

## General

What is the scale? - 1 Pixel corresponds to about 390 nm.

We will first look at the spatial data from all samples in one plot. The
cell types are annotated according to the information the authors of the 
original study provided.

As pointed out in their publication, one can see that there is a continuum 
between mixing of immune cells (yellow) and cancer tissue (green and blue).
Either we can see a clear interface between both populations or they are
totally mixed. This is actually associated with survival (see paper).
Below this observation is quantified by calculating a similar mixing score
as they did in the original publication.

```{r fig.height=12, fig.width=18}
meta %>%
  ggplot() +
  geom_point(aes(x = row, y = col, col = Group), size = .1) +
  facet_wrap(~ SampleID) + 
  guides(colour = guide_legend(override.aes = list(size = 4)))
```

## Mixing Score

Calculating the mixing score by counting for all immune cells in each
sample how many contacts to they have to cancer cells divided by the 
contacts to other immune cells. The contacts are determined by
Delaunay Triangulation and cells are only considered to be neighbors
if the distance between them is below 40. 

Importantly this mixing score is higher if there are more tumor cells
than immune cells (as seen by the filling color of the barplot below,
`ratio.count.tumor.immune` refers to the number of tumor cells 
divided by the number of immune cells).

(Remark: The authors of the paper used Haralick’s gray-level co-occurrence
matrix to determine direct neighbors of each cell)

```{r}
# Setting the distance threshold for what is considered a neighbor in the DT.
neighbor.thr <- 40

if ("tumor.immune.contacts.RDS" %in% list.files("data")) {
  contacts.smp <- readRDS("data/tumor.immune.contacts.RDS")
} else {
  
  coord.type.smp <- unique(meta$SampleID) %>%
  set_names(paste0("s", .)) %>%
  map(function(id) {
    meta %>%
      filter(meta$SampleID == id) %>% 
      select(row, col, Group) %>%
      mutate(Group = ifelse(Group == "Keratin-positive tumor", 
                            "Tumor", Group)) %>%
      filter(Group %in% c("Tumor", "Immune")) %>%
      mutate(id = seq_len(nrow(.)))
  })

  # helper function for getting direct neighbors
  get_neighbors <- function(ddobj, id) {
    dplyr::union(
      ddobj$delsgs$ind1[which(ddobj$delsgs$ind2 == id)],
      ddobj$delsgs$ind2[which(ddobj$delsgs$ind1 == id)]
    )
  }

  contacts.smp <- coord.type.smp %>%
  set_names() %>%
  map_dfr(function(smp) {
    
    delaunay <- deldir::deldir(as.data.frame(smp %>% select(row, col)))
    dists <- distances::distances(as.data.frame(smp %>% select(row, col)))
    ids.to.map <- smp %>% filter(Group == "Immune") %>% pull(id)
    
    map(ids.to.map, function(i) {
      alln <- get_neighbors(delaunay, i)
      actualn <- alln[which(dists[alln, i] <= neighbor.thr)]
      
      smp %>%
        filter(id %in% actualn) %>%
        pull(Group)
      
    }) %>% 
      unlist %>%
      table
  })
}

meta %>%
  group_by(SampleID) %>%
  count(Group) %>% 
  pivot_wider(names_from = Group, values_from=n, names_prefix = "Count.") %>%
  cbind(contacts.smp) %>%
  mutate(ratio.tumor.immune = as.vector(Tumor/Immune * 100)) %>%
  rename("Count.Keratin.positive.Tumor" = "Count.Keratin-positive tumor") %>%
  replace(is.na(.), 0) %>%
  mutate(Count.Tumor.Total = Count.Keratin.positive.Tumor + Count.Tumor) %>%
  mutate(ratio.count.tumor.immune = Count.Tumor.Total / Count.Immune) %>%
  ggplot() +
  geom_bar(aes(x=reorder(SampleID, -ratio.tumor.immune), 
               y=ratio.tumor.immune,
               fill=ratio.count.tumor.immune), 
           stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Sample", y = "Mixing Score") +
  scale_fill_viridis_c()
```

Looking at the samples with the top mixing score, namely 15, 18, 22, 24, 31, 
25, 21, 26

```{r fig.width=12, fig.height=6}
meta %>%
  filter(SampleID %in% c(15, 18, 22, 24, 31, 25, 21, 26)) %>%
  ggplot() +
  geom_point(aes(x = row, y = col, col = Group), size = .1) +
  facet_wrap(~ SampleID, ncol = 4) + 
  guides(colour = guide_legend(override.aes = list(size = 4)))
```

Looking at the samples with the lowest mixing score, namely 35, 28, 16, 37, 
4, 40, 41, 5.

```{r fig.width=12, fig.height=6}
meta %>%
  filter(SampleID %in% c(35, 28, 16, 37, 4, 40, 41, 5)) %>%
  ggplot() +
  geom_point(aes(x = row, y = col, col = Group), size = .1) +
  facet_wrap(~ SampleID, ncol = 4) + 
  guides(colour = guide_legend(override.aes = list(size = 4)))
```

And we can see that those mixing scores resemble exactly what we see
in the images.

## Sample 40

Looking more closely at sample 40 to make sure that the preprocessing
went right. First, we will look at the cells only using their
calculated center of mass.

```{r fig.height=6, fig.width=8}
meta %>%
  filter(SampleID == 40) %>%
  ggplot() +
  geom_point(aes(x = row, y = col, col = Group), size=.4)  + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
```

Furthermore, we will look at the tiff.

```{r fig.height=6, fig.width=8}
library(tiff)
id = 40
tiff <- readTIFF(paste0("data/raw_tiffs/p", 
                        id, "_labeledcellData.tiff"), as.is = TRUE)
```

First, the binary tiff, meaning all values that there not 0 or 1 (i.e. a
cell was detected).

```{r fig.height=6, fig.width=7}
# get coordinates for where there was a cell detected (=: binary)
binary <- apply(tiff, c(1,2), function(n) !(n %in% c(0,1)))
binary.df <- as.data.frame(binary)
rownames(binary.df) <- seq_len(nrow(binary))
colnames(binary.df) <- seq_len(ncol(binary))

df <- binary.df %>% 
  rownames_to_column(var = "row") %>%
  pivot_longer(cols = !(row), names_to = "col") %>%
  mutate(row = as.numeric(row), col = as.numeric(col))

subs <- sample(nrow(df), .03 * nrow(df))

df %>%
  slice(n = subs) %>%
  mutate(cell.present = value) %>%
  ggplot() +
  geom_point(aes(x=row, y=col, col=cell.present), size=.1) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()) +
  scale_colour_manual(values = c("TRUE" = "green2", 
                                 "FALSE" = "blue4"))  + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
```

Second, we will visualize the tiff and the labeled center of mass together.
And on this scale the result actually looks quite nice.

```{r fig.height=6, fig.width=8}
s.40 <- meta %>%
  filter(SampleID == 40)

tiff.40 <- df %>%
  filter(value) %>%
  slice_sample(., n = nrow(.)/10)

ggplot() +
  geom_point(data = tiff.40, aes(x=row, y=col), col="black", 
             size = 0.1, stroke = 0, shape = 16) +
  geom_point(data = s.40, aes(x=row, y=col, col=Group), size = .5) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank())  + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
```

To qualitatively assess the performance of the segmentation we will furthermore
look at a close up of sample 40, more specifically the region from row 0 to
row 600 and from col 1448 to col 2048. And we can see
that the center of mass (colored point) mostly matches the cell boundaries,
but not always!

```{r fig.height=8, fig.width=10}
c.max <- dim(binary)[1]
top_left <- list(row = 0, col = c.max)
bottom_right <- list(row = 600, col = c.max-600)

# get the unique coordinates for each cell
tiff.df <- as.data.frame(tiff)
rownames(tiff.df) <- seq_len(nrow(tiff))
colnames(tiff.df) <- seq_len(ncol(tiff))

tiff.df.plot <- tiff.df %>%
  rownames_to_column(var = "row") %>%
  pivot_longer(cols = !(row), names_to = "col") %>%
  mutate(row = as.numeric(row), col = as.numeric(col)) %>%
  filter(!(value %in% c(0, 1))) %>%
  # how to get only entry for each unique value
  group_by(value) %>%
  filter(row_number()==1) %>%
  ungroup()

s.40 <- meta %>%
  filter(SampleID == 40) %>%
  filter(row >= top_left$row, row <= bottom_right$row,
         col >= bottom_right$col, col <= top_left$col)

tiff.40 <- df %>%
  filter(value) %>%
  filter(row >= top_left$row, row <= bottom_right$row,
         col >= bottom_right$col, col <= top_left$col)

labels.40 <- tiff.df.plot %>%
  filter(row >= top_left$row, row <= bottom_right$row,
         col >= bottom_right$col, col <= top_left$col)

ggplot() +
  geom_point(data = tiff.40, aes(x=row, y=col), col="black", 
             size = 0.1, stroke = 0, shape = 16) +
  geom_point(data = s.40, aes(x=row, y=col, col=Group), size = 3) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank())
```

Here are the corresponding IDs for comparison to the meta data.

```{r fig.height=7, fig.width=8}
ggplot(labels.40) +
  geom_label(aes(x = row, y = col, label = value), size = 2.5)
```

# Merging Metadata

Merging the meta data with the mixing score and some other measures.

```{r}
# Generating Sample Meta Data
immune.counts <- meta %>%
  group_by(SampleID) %>%
  count(Group) %>%
  pivot_wider(names_from = Group, values_from = n,
              names_prefix = "Count.")

total.counts <- meta %>%
  count(SampleID) %>%
  rename(Count.Total = n)

mixing.score <- round(
  contacts.smp %>% 
    mutate(ratio.tumor.immune = Tumor/Immune * 100) %>%
    pull(ratio.tumor.immune),
  2)

sample.meta <- patient.info %>%
  filter(SampleID != 30) %>%
  mutate(ratio.tumor.immune = as.vector(mixing.score)) %>%
  inner_join(immune.counts, by = "SampleID") %>%
  inner_join(total.counts, by = "SampleID") %>%
  # see paper Figure 3, Panel C
  mutate(Classification = 
         case_when(Classification == 0 ~ "mixed",
                   Classification == 1 ~ "compartimentalized",
                   Classification == 2 ~ "cold")) %>% 
  mutate(SampleID = as.character(SampleID)) %>%
  left_join(patient.info.paper, by = c("SampleID" = "InternalId"))

sample.meta %>% slice_head(n=6)
```

# Running MISTy

## Preparation

Select only proteins from the expression matrix and check for NAs.

Also we need to check the variance for each marker in each same and
exclude the marker if the variance is 0. The same is true for markers
whose number of unique values is lower than the desired number of cv
folds (here 10).

```{r collapse=TRUE}
cv.folds = 10
expr.smp <- unique(meta$SampleID) %>%
  set_names(paste0("s", .)) %>%
  map(function(id) {
    ret.expr <- expr %>%
      filter(meta$SampleID == id) %>%
      # Select only proteins
      select(11:47)
    
    # Check for zero variance (otherwise MISTy throws an error)
    target.var <- apply(ret.expr, 2, stats::sd, na.rm = TRUE)
    ret.expr <- ret.expr %>% select(-names(which(target.var == 0)))
    
    # Check for how many unique values
    target.unique <- colnames(ret.expr) %>%
      purrr::set_names() %>%
      purrr::map_int(~ length(unique(ret.expr %>% pull(.x))))
    ret.expr <- ret.expr %>% select(
      names(target.unique[target.unique > cv.folds])
      )
      
    colnames(ret.expr) <- make.names(colnames(ret.expr))
    ret.expr
  })

# Check NAs for sample 40
sum(map_int(names(expr.smp$s40), ~ sum(is.na(expr.smp$s40[.x]))))

coord.smp <- unique(meta$SampleID) %>%
  set_names(paste0("s", .)) %>%
  map(function(id) {
    meta %>%
      filter(meta$SampleID == id) %>% 
      select(row, col)
  })

# Check NAs for sample 40
sum(map_int(names(coord.smp$s40), ~ sum(is.na(coord.smp$s40[.x]))))

nrow(expr.smp$s40) == nrow(coord.smp$s40)
```

Helper function to calculate distances to the neighbors.

```{r}
# helper function for nearest neighbors
dist_k_nn <- function(sample, k) {
  nn <- RANN::nn2(sample, eps=0.0, k = k)
  map_dbl(seq_len(nrow(sample)), function(n) {
    mean(nn$nn.dists[n, ])
  })
}

# helper function for getting direct neighbors
get_neighbors <- function(ddobj, id) {
  dplyr::union(
    ddobj$delsgs$ind1[which(ddobj$delsgs$ind2 == id)],
    ddobj$delsgs$ind2[which(ddobj$delsgs$ind1 == id)]
  )
}
```

First we will look at the distribution of the  mean distances to the direct 
neighbors (as determined by the delauney triangulation) to get a feeling for 
which cutoff value should be used for the generation of the juxtaview. The
corresponding distribution is shown below.
(in blue the cutoff - 40 pixels (15.6 microns) - which was used in the end).

```{r}
# mean distance to the direct neighbors for each cell
mean.dists.direct.neighbors <- map(coord.smp, function(smp) {
  delaunay <- deldir::deldir(as.data.frame(smp))
  dists <- distances::distances(as.data.frame(smp))
  map_dbl(seq_len(nrow(smp)), function(id) {
    nn <- get_neighbors(delaunay, id)
    mean(dists[nn, id])
  })
}) 

stat.per.sample <- map2_dfr(mean.dists.direct.neighbors,
                            seq_len(length(mean.dists.direct.neighbors)),
                            function(list, smp) {
            data.frame(sample = smp, value = list)
          })

# histogram
ggplot(data.frame(dist = mean.dists.direct.neighbors %>% unlist)) +
  geom_histogram(aes(dist), bins = 100) +
  lims(x = c(0,100)) +
  geom_vline(xintercept = 40, col = "blue")
```

Summary statistics:

```{r}
data.frame(value = mean.dists.direct.neighbors %>% unlist) %>%
  summarise(mean = mean(value), 
            median = median(value), 
            q25 = quantile(value, 0.25),
            q75 = quantile(value, 0.75),
            min = min(value),
            max = max(value))
```

However, a more detailed look across the sample (boxplot) shows that the
cutoff may be too low for certain samples. It may be good to increase this
threshold to 50.

```{r}
# boxplot
stat.per.sample %>%
  mutate(sample = factor(sample)) %>%
  ggplot() +
  geom_boxplot(aes(x=sample, y=value)) +
  geom_hline(yintercept = 40, col="blue") +
  lims(y = c(0, 200))
```

Furthermore we will look at the distribution of the mean distances to the
10 -, 20 -, 40 -, and 80 closest cells to get a feeling
for which cutoff will be used to generate the paraview 
(in blue the cutoff - 120 pixels (46.8 microns) - which was used in the end).

```{r}
neighbor.dists <- map_dfr(coord.smp, function(smp) {
  c(10, 20, 40, 80) %>%
    set_names() %>%
    map_dfc(function(k) {
  dist_k_nn(smp, k)
    }) %>%
  set_names(paste0("dist_to_", c("10", "20", "40", "80")))
})

neighbor.dists %>%
  pivot_longer(cols = everything()) %>%
  mutate(name = factor(name, 
                       levels = paste0("dist_to_", c("10", "20", "40", "80")))) %>%
  ggplot() +
  geom_histogram(aes(value), bins = 100) +
  facet_wrap(~ name) +
  lims(x = c(0, 200)) +
  geom_vline(xintercept = 120, col = "blue")
```

Summary statistics:

```{r}
neighbor.dists %>%
  pivot_longer(cols = everything()) %>%
  group_by(name) %>%
  summarise(mean = mean(value), 
            median = median(value), 
            q25 = quantile(value, 0.25),
            q75 = quantile(value, 0.75),
            min = min(value),
            max = max(value))
```

## Generating the MISTy views

Based on the data above, the following parameters (thresholds) were used 
to run MISTy: 
- Juxtaview: threshold 40,
- Paraview: theshold 120, zone of indifference: 40

```{r}
if ("misty_views_smp.RDS" %in% list.files("data")) {
  misty.views.smp <- readRDS("data/misty_views_smp.RDS")
} else {
  misty.views.smp <- map(expr.smp, coord.smp, function(expr, coord) {
    
    # Create views and run MISTy
    create_initial_view(expr) %>%
      add_juxtaview(positions = coord, neighbor.thr = 40) %>%
      add_paraview(positions = coord, l = 120, zoi = 40)
  })
  names(misty.views.smp) <- names(expr.smp)
  saveRDS(misty.views.smp, "data/misty_views_smp.RDS")
}
```

## Random Forest Implementation

Running with random forest as ML algorithm to build the view-specific models. 

```{r}
if ("ranger.results.RDS" %in% list.files("data")) {
  ranger.results <- readRDS("data/ranger.results.RDS")
} else {
  cv.folds = 10
  ranger.results.folders <- map2(
    misty.views.smp, names(misty.views.smp), function(smp, name) {
      
      smp %>% run_misty(results.folder = paste0("analysis/results/ranger_", name),
                        cv.folds = cv.folds)
    })
  ranger.results <- collect_results(ranger.results.folders)
  saveRDS(ranger.results, "data/ranger.results.RDS")
}
```

# MISTy Results

## Gain R2

We can see that there is quite some gain in variance when taking the 
expression of neighboring cells into account. However, looking at the
errorbars we can also see that there is a large variability between the samples.

```{r}
mistyR::plot_improvement_stats(ranger.results)
```

Let's look at a heatmap first to see how the gain.R2 stat is distributed across
the samples and targets.

```{r fig.width=8, fig.height=7}
ranger.results$improvements %>% 
  filter(measure == "gain.R2") %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  select(-measure) %>% 
  pivot_wider(names_from = target) %>%
  replace(is.na(.), 0) %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample") %>%
  as.matrix() %>%
  pheatmap::pheatmap()
```

```{r include=FALSE, eval=FALSE}
# see tutorial here:
# https://www.datacamp.com/community/tutorials/hierarchical-clustering-R
```

We can see in the dendogram that there is some structure in the samples 
regarding their gain.R2 signature. But what is driving this structure?
In the following we are going to classify the cells based on the 
hierachical clustering and compare the properties of these clusters.

We will take 10 clusters but remove all clusters that have only one sample in them.
This results in 7 clusters overall.

```{r}
k = 10

hc <- ranger.results$improvements %>% 
  filter(measure == "gain.R2") %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  select(-measure) %>% 
  pivot_wider(names_from = target) %>%
  replace(is.na(.), 0) %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample") %>%
  as.matrix() %>%
  dist() %>%
  hclust()

plot(hc)
rect.hclust(hc , k = k, border = 2:6)
```

Removing clusters with only 1 sample in them.

```{r}
hc.clusters <- cutree(hc, k = k)

n <- unique(hc.clusters) %>%
  set_names() %>%
  map_int(function(i) {
    s <- sum(hc.clusters == i)
  })
keep <- names(n)[which(n > 1)]
n
```

Now visualize the heatmap again with the clusters.

```{r fig.width=8, fig.height=7}
col.annotation <- data.frame(cluster = factor(hc.clusters))
rownames(col.annotation) <- names(hc.clusters)

ranger.results$improvements %>% 
  filter(measure == "gain.R2") %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(tibble(sample = names(hc.clusters), cluster = hc.clusters), 
            by = "sample") %>%
  filter(cluster %in% keep) %>%
  select(-c(measure, cluster)) %>% 
  pivot_wider(names_from = target) %>%
  replace(is.na(.), 0) %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample") %>%
  as.matrix() %>%
  pheatmap::pheatmap(., annotation_row = col.annotation)
```

### What differentiates the clusters? {.tabset}

#### Cell Counts and Mixing Score

Could these quantitaive properties explain the differences:
* ratio.tumor.immune
* Count.Immune
* Count.Tumor
* Count.Endothelial
* Count.Mesenchymal.like
* Count.Total

One can immediately see that cluster 2 has a higher ration of tumor-immune to
immune-immune contact, though this effect seems to be mainly driven by the
cold tumors (low number of immune cells).

In general cluster 2 contains the samples with low total cell numbers and
with a high number of tumor cells compared to immune cells.

```{r fig.height=5, fig.width=8}
sample.meta %>%
  left_join(tibble(sample = names(hc.clusters), cluster = factor(hc.clusters)), 
            by = c("SampleID" = "sample")) %>%
  filter(cluster %in% keep) %>%
  rename("Count.Keratin.positive.Tumor" = "Count.Keratin-positive tumor",
                       "Count.Mesenchymal.like" = "Count.Mesenchymal-like") %>%
  mutate(Count.Tumor = Count.Tumor + Count.Keratin.positive.Tumor) %>%
  pivot_longer(cols = c(ratio.tumor.immune, Count.Immune, Count.Tumor, 
                        Count.Endothelial, Count.Mesenchymal.like, Count.Total),
               names_to = "statistic", values_to = "value") %>%
  ggplot() +
  geom_boxplot(aes(x = cluster, y = value), outlier.shape=NA) +
  geom_jitter(aes(x = cluster, y = value, col = Classification), size=.8, width=.3) +
  facet_wrap(~ statistic, scales = "free", ncol = 3) +
  theme(legend.position="top") +
  scale_color_manual(values = c("mixed" = "indianred", 
                               "compartimentalized" = "forestgreen",
                               "cold" = "royalblue"))
```

Though it may be more informative to actually look at the cell ratios, or 
proportions. 

#### Spatial Distribution

What about the spatial distribution of the cells? Could it be a contributing 
factor? Let's look at the mean distances to the neighboring cells 
(as determined by Delaunay Triangulation) for each cell in a given cluster. 
And it seems like the distances are larger cluster 2 compared to the rest.

```{r}
stat.per.sample %>%
  mutate(sample = as.character(sample)) %>%
  inner_join(tibble(sample = names(hc.clusters), cluster = factor(hc.clusters)), 
            by = "sample") %>%
  filter(cluster %in% keep) %>%
  ggplot() +
  geom_boxplot(aes(x=cluster, y=value)) +
  lims(y = c(0, 100))
```

#### Classification

What about the proportion of the classification 
(referring to mixed, compartmentalized, or cold) accross the clusters?

```{r fig.height=4, fig.width=8}
p1 <- sample.meta %>%
  left_join(tibble(sample = names(hc.clusters), cluster = factor(hc.clusters)), 
            by = c("SampleID" = "sample")) %>%
  filter(cluster %in% keep) %>%
  rename("Count.Keratin.positive.Tumor" = "Count.Keratin-positive tumor",
                       "Count.Mesenchymal.like" = "Count.Mesenchymal-like") %>%
  mutate(Count.Tumor = Count.Tumor + Count.Keratin.positive.Tumor) %>%
  group_by(cluster) %>%
  summarise(n = n(), 
         n.mixed = sum(Classification == "mixed"),
         n.comp = sum(Classification == "compartimentalized"),
         n.cold = sum(Classification == "cold")) %>%
  mutate(fraction.mixed = n.mixed/n,
         fraction.comp = n.comp/n,
         fraction.cold = n.cold/n) %>%
  pivot_longer(cols = starts_with("frac")) %>%
  ggplot() +
  geom_col(aes(x = cluster, y = value, fill = name)) +
  geom_text(aes(x = cluster, y = 0.99, label = n), vjust = -0.5) +
  scale_fill_manual(values = c("fraction.mixed" = "indianred", 
                               "fraction.comp" = "forestgreen",
                               "fraction.cold" = "royalblue")) +
  labs(title = "Fractions")

p2 <- sample.meta %>%
  left_join(tibble(sample = names(hc.clusters), cluster = factor(hc.clusters)), 
            by = c("SampleID" = "sample")) %>%
  filter(cluster %in% keep) %>%
  rename("Count.Keratin.positive.Tumor" = "Count.Keratin-positive tumor",
                       "Count.Mesenchymal.like" = "Count.Mesenchymal-like") %>%
  mutate(Count.Tumor = Count.Tumor + Count.Keratin.positive.Tumor) %>%
  group_by(cluster) %>%
  summarise(n = n(), 
         n.mixed = sum(Classification == "mixed"),
         n.comp = sum(Classification == "compartimentalized"),
         n.cold = sum(Classification == "cold")) %>%
  pivot_longer(cols = starts_with("n.")) %>%
  ggplot() +
  geom_col(aes(x = cluster, y = value, fill = name)) +
  scale_fill_manual(values = c("n.mixed" = "indianred", 
                               "n.comp" = "forestgreen",
                               "n.cold" = "royalblue")) +
  labs(title = "Absolute Numbers")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

#### Intra and Multiview

Could it also be possible that the the intraview R2 somehow explains the clustering
we see in the gain.R2. Meaning samples in which the intraview R2s are
rather low and samples in which it is rather high?

First, we one can note that for the targets there seem to be three
distinct clusters. On the left the intra R2 is rather high, in the
middle the intra R2 is rather low, and on the right the intra R2 
values are mixed across the samples.

Regarding the question, the clusters seem to be somewhat preserved, so the
intraview R2 probably has something to do with the observed clusterin in gain R2
(the intraview basically determines the potential of gaining explained variance).

```{r fig.width=8, fig.height=7}
ranger.results$improvements %>% 
  filter(measure == "intra.R2") %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(tibble(sample = names(hc.clusters), cluster = hc.clusters), 
            by = "sample") %>%
  filter(cluster %in% keep) %>%
  select(-c(measure, cluster)) %>% 
  pivot_wider(names_from = target) %>%
  replace(is.na(.), 0) %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample") %>%
  as.matrix() %>%
  pheatmap::pheatmap(., annotation_row = col.annotation)
```

How does the multiview R2 heatmap look like?

Regarding the sample clusters one can see quite clearly that the clusters
from the gain.R2 statistic are not fully preserved. For a given sample
the neighboring samples often belong to the same cluster, but the clusters
seem to be broken into several smaller subsets (e.g. look at clusters 2 and 4).

```{r fig.width=8, fig.height=7}
ranger.results$improvements %>% 
  filter(measure == "multi.R2") %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(tibble(sample = names(hc.clusters), cluster = hc.clusters), 
            by = "sample") %>%
  filter(cluster %in% keep) %>%
  select(-c(measure, cluster)) %>% 
  pivot_wider(names_from = target) %>%
  replace(is.na(.), 0) %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample") %>%
  as.matrix() %>%
  pheatmap::pheatmap(., annotation_row = col.annotation)
```

Another way to go about this investigation may be to select the top targets
in terms of gain R2 and then check what this gain R2 is correlated with
The details can be found below.

#### From the Targets

Let's take the 12 targets (1/3 of all targets) for which the mean gain.R2 is the largest and
investigate the differences between the samples.

```{r}
top.targets <- ranger.results$improvements %>%
  filter(measure == "gain.R2") %>%
  group_by(target) %>%
  summarise(mean.gain.R2 = mean(value), .groups = "drop") %>%
  slice_max(mean.gain.R2, n=12) %>%
  pull(target)
top.targets
```

How should the summarise the gained explained variance per sample? Sum it up, 
take the mean, median?

```{r fig.width=8, fig.height=8}
plots <- map2(c(mean, median, sum), c("Mean", "Median", "Sum"), function(func, name) {
  ranger.results$improvements %>%
  filter(measure == "gain.R2") %>% 
  filter(target %in% top.targets) %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  group_by(sample) %>%
  summarise(stat = func(value), .groups = "keep") %>%
  ggplot() +
  geom_bar(aes(x = reorder(sample, -stat), y = stat), 
           stat = "identity") +
  labs(title = name)
})

do.call(gridExtra::grid.arrange, c(plots, ncol=1))
```

To know which statistic might be useful, we should figure out
how our gain.R2 stat for each sample is dependent on the targets.

```{r}
ordering <- ranger.results$improvements %>% 
  filter(measure == "gain.R2") %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  group_by(sample) %>%
  summarise(m = median(value)) %>%
  arrange(desc(m)) %>%
  pull(sample)

ranger.results$improvements %>%
  filter(measure == "gain.R2") %>% 
  filter(target %in% top.targets) %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  mutate(sample = factor(sample, levels = ordering)) %>%
  ggplot() +
  geom_boxplot(aes(x = sample, y = value))
```

Let's look at some correlation between other sample features and the
gained explained variance (R2).

Quantiative Predictors:
- Mixing score
- Count of immune cells
- Count of tumor cells
- Count of endothelial cells
- Count of mesenchymal-like cells
- Total cell count

```{r fig.height=18, fig.width=10}
plots <- map(c("ratio.tumor.immune", "Count.Immune", "Count.Tumor", 
               "Count.Endothelial", "Count.Mesenchymal.like", "Count.Total"),
              function(test.var) {
  ranger.results$improvements %>%
    filter(measure == "gain.R2") %>% 
    filter(target %in% top.targets) %>%
    mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
    group_by(sample) %>%
    summarise(sum.gain.R2 = sum(value), mean.gain.R2 = mean(value),
              median.gain.R2 = median(value), .groups = "drop") %>%
    left_join(sample.meta %>%
                rename("Count.Keratin.positive.Tumor" = "Count.Keratin-positive tumor",
                       "Count.Mesenchymal.like" = "Count.Mesenchymal-like") %>%
                mutate(Count.Tumor = Count.Tumor + Count.Keratin.positive.Tumor),
              by = c("sample" = "SampleID")) %>%
    pivot_longer(cols = c(sum.gain.R2, mean.gain.R2, median.gain.R2), 
                 names_to = "stat") %>%
    mutate(stat = factor(stat, levels = c("mean.gain.R2", "sum.gain.R2", "median.gain.R2"))) %>%
    ggplot() +
    geom_point(aes(x = !!sym(test.var), y = value)) +
    geom_smooth(aes(x = !!sym(test.var), y = value),
                formula = y ~ x, method = "lm") +
    facet_wrap(~ stat, ncol=3, scales="free_y") +
    labs(title = test.var)
})

do.call(gridExtra::grid.arrange, c(plots, ncol=1))
```

Correlation Plot:

```{r}
walk(c("sum.gain.R2", "mean.gain.R2", "median.gain.R2"), function(test.var) {
  ranger.results$improvements %>%
    filter(measure == "gain.R2") %>% 
    filter(target %in% top.targets) %>%
    mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
    group_by(sample) %>%
    summarise(sum.gain.R2 = sum(value), mean.gain.R2 = mean(value),
              median.gain.R2 = median(value), .groups = "drop") %>%
    left_join(sample.meta %>%
                rename("Count.Keratin.positive.Tumor" = "Count.Keratin-positive tumor",
                       "Count.Mesenchymal.like" = "Count.Mesenchymal-like") %>%
                select(1:10) %>%
                replace(is.na(.), 0) %>%
                mutate(Count.Tumor = Count.Tumor + Count.Keratin.positive.Tumor),
              by = c("sample" = "SampleID")) %>%
    select(!!sym(test.var), ratio.tumor.immune,
           Count.Immune, Count.Tumor, Count.Endothelial, Count.Mesenchymal.like,
           Count.Total) %>%
    cor() %>%
    corrplot::corrplot(type = "upper", tl.col = "black")
})
```

Now taking the top 10 and the bottom 10 samples and comparing their features.
Features being the mixing score, the R2 of the intraview for the chosen targets

```{r}
top.10 <- ranger.results$improvements %>%
  filter(measure == "gain.R2") %>% 
  filter(target %in% top.targets) %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  group_by(sample) %>%
  summarise(mean.gain.R2 = sum(value), .groups = "drop") %>%
  slice_max(mean.gain.R2, n = 10) %>%
  pull(sample)

bottom.10 <- ranger.results$improvements %>%
  filter(measure == "gain.R2") %>% 
  filter(target %in% top.targets) %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  group_by(sample) %>%
  summarise(mean.gain.R2 = sum(value), .groups = "drop") %>%
  slice_min(mean.gain.R2, n = 10) %>%
  pull(sample)
```

Looking at the mixing score (ratio of immune-tumor contacts divided by the 
number of immune-immune contacts). And we can kinda see that the 

```{r}
sample.meta %>%
  mutate(group.tmp = case_when(
    SampleID %in% top.10 ~ "top",
    SampleID %in% bottom.10 ~ "bottom",
    TRUE ~ "middle",
  )) %>%
  mutate(group.tmp = factor(group.tmp, levels = c("top", "middle", "bottom"))) %>%
  ggplot() +
  geom_boxplot(aes(x=group.tmp, y=ratio.tumor.immune))
```

## View Contributions

```{r}
mistyR::plot_view_contributions(ranger.results)
```

Also looking at the variability.

```{r fig.width=6, fig.height=6}
ranger.results$contributions %>%
  filter(!str_starts(view, "p\\."), view != "intercept") %>%
  ggplot() +
  geom_boxplot(aes(x = target, y = value)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_wrap(~ view, ncol = 1)
```

## Importances

```{r fig.width=6, fig.height=6}
mistyR::plot_interaction_heatmap(ranger.results, view = "intra", clean = TRUE)
mistyR::plot_interaction_heatmap(ranger.results, view = "juxta.40", clean = TRUE)
mistyR::plot_interaction_heatmap(ranger.results, view = "para.120", clean = TRUE)
```

## Signatures

### Performance

```{r}
ranger.perf.sig <- mistyR::extract_signature(ranger.results, "performance")
ranger.perf.pca <- stats::prcomp(ranger.perf.sig %>% select(-sample))
```

```{r fig.width=16, fig.height=10}
p1 <- ggplot() +
  geom_point(aes(x=ranger.perf.pca$x[,1], y=ranger.perf.pca$x[,2],
                 col=factor(sample.meta$GRADE))) +
  labs(title = "Grade")

p2 <- ggplot() +
  geom_point(aes(x=ranger.perf.pca$x[,1], y=ranger.perf.pca$x[,2],
                 col=factor(sample.meta$Classification))) +
  labs(title = "Classification")

p3 <- ggplot() +
  geom_point(aes(x=ranger.perf.pca$x[,1], 
                 y=ranger.perf.pca$x[,2],
                 col=log10(sample.meta$ratio.tumor.immune))) +
  scale_color_viridis_c() +
  labs(title = "Mixing Score")

p4 <- ggplot() +
  geom_point(aes(x=ranger.perf.pca$x[,1], 
                 y=ranger.perf.pca$x[,2],
                 col=sample.meta$Count.Total)) +
  scale_color_viridis_c() +
  labs(title = "Total Cell Counts")

gridExtra::grid.arrange(p1, p2, p3, p4, ncol=2)
```

Which factors drive the variance in the results?

```{r}
fviz_pca_var(ranger.perf.pca, col.var="contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

### Importance

```{r}
ranger.imp.sig <- mistyR::extract_signature(ranger.results, "importance")
ranger.imp.pca <- stats::prcomp(ranger.imp.sig %>% select(-sample))
```

Color by Grade.

```{r fig.width=16, fig.height=5}
p1 <- ggplot() +
  geom_point(aes(x=ranger.imp.pca$x[,1], y=ranger.imp.pca$x[,2],
                 col=factor(sample.meta$GRADE))) +
  labs(title = "Grade")

p2 <- ggplot() +
  geom_point(aes(x=ranger.imp.pca$x[,1], y=ranger.imp.pca$x[,2],
                 col=factor(sample.meta$Classification))) +
  labs(title = "Classification")

gridExtra::grid.arrange(p1, p2, ncol=2)
```

Looking at the variable loadings one can see that the importances are often 
similar for different views and that the importances are reciprocal, e.g.
- CD45 & HLA_Class_1 (= MHC1)
- CD63 & CD68 (macrophages)
- CD16 & CD11c (natural killer cells, neutrophils, monocytes, and macrophages)
- CD16 & CD45

```{r}
fviz_pca_var(ranger.imp.pca, col.var="contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

# Save data

```{r}
saveRDS(expr, "data/expression.RDS")
saveRDS(expr.smp, "data/expression_per_sample.RDS")
saveRDS(meta, "data/meta_data.RDS")
saveRDS(sample.meta, "data/sample_meta.RDS")
```

