---
title: "Cell Type-specific Analysis"
---

# Set up

Messages will be hidden and the output collapsed to make the Rmd more clear.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = "~/Saez/workflowr_projects/mibi_analysis/")
```

Loaded packages.

```{r}
library(mistyR)
library(future)
plan("multisession", workers=14)
library(tidyverse)
library(factoextra)
```

Loading data generated in MIBI_Analysis.Rmd

```{r}
all.expr <- readRDS("data/expression.RDS")
meta <- readRDS("data/meta_data.RDS")
meta.smp <- readRDS("data/sample_meta.RDS")
```

# Rationale

I think running MISTy on stratified samples does not make sense since we can
easily stratify the results later on based on certain features such as
tumor grade, tumor classification, etc. and whether we include a certain
sample does not change the MISTy results for another sample.

Therefore, stratification makes more sense on the level of units (here 
single cells) for example by running MISTy only on immune cells and using
only cancer cells to create the paraview. This way we could try to find out
which tumor markers influence the immune cell/response.

# Computing MISTy views

## General Parameters

```{r}
cv.folds <- 10
```

## Tumor -> Immune

Import the get_weight function.

```{r}
# only removed the constant part here.
get_weight <- function(family = c("gaussian", "exponential", "linear"),
                       distances, parameter, zoi) {
  expr.family <- match.arg(family)
  
  distances[distances < zoi] <- Inf
  dim.orig <- dim(distances)
  
  switch(expr.family,
    "gaussian" = {
      exp(-distances^2 / parameter^2)
    },
    "exponential" = {
      exp(-distances / parameter)
    },
    "linear" = {
      weights <- pmax(0, 1 - distances / parameter)
      dim(weights) <- dim.orig
      weights
    }
  )
}
```

Show how the cancer -> immune views are genereated for the first sample.

```{r}
# Get expression, positions, and types for a sample
expr <- all.expr[meta$SampleID == 1, ] %>% select(11:47)
pos <- meta %>% filter(SampleID == 1) %>% select(row, col)
types <- meta %>% filter(SampleID == 1) %>% pull(Group)

# Get the ids of immune cells
immune.ids <- which(types == "Immune")

# Get the ids of cancer cells
cancer.ids <- which(types %in% c("Keratin-positive tumor", "Tumor"))

# Get the distances between all cells
dists <- distances::distances(as.data.frame(pos))

# Parameters for get weight
l <- 120
zoi <- 40
family <- "gaussian"

view.df <- map_dfr(immune.ids, function(i) {
  weights <- get_weight(family, dists[i, ], l, zoi)
  weighted.sum <- colSums(expr[cancer.ids, ] * weights[cancer.ids])
  data.frame(t(weighted.sum))
})

# Sanity Check for view.df

# 1. Check for zero variance (otherwise MISTy throws an error)
target.var <- apply(view.df, 2, stats::sd, na.rm = TRUE)
view.df <- view.df %>% select(-names(which(target.var == 0)))

# 2. Check for how many unique values
target.unique <- colnames(view.df) %>%
  purrr::set_names() %>%
  purrr::map_int(~ length(unique(view.df %>% pull(.x))))

view.df <- view.df %>% select(
  names(target.unique[target.unique > cv.folds])
  )

view <- create_view(paste0("cancer.paraview.", l),
            view.df,
            paste0("c.para.", l))

# Sanity Check for intra.df
# 1. Check for zero variance (otherwise MISTy throws an error)
intra.df <- expr[immune.ids,]
target.var <- apply(intra.df, 2, stats::sd, na.rm = TRUE)
intra.df <- intra.df %>% select(-names(which(target.var == 0)))

# 2. Check for how many unique values
target.unique <- colnames(intra.df) %>%
  purrr::set_names() %>%
  purrr::map_int(~ length(unique(intra.df %>% pull(.x))))

intra.df <- intra.df %>% select(
  names(target.unique[target.unique > cv.folds])
  )

# Make correct colnames
colnames(intra.df) <- make.names(colnames(intra.df))
colnames(view.df) <- make.names(colnames(view.df))

# Create custom "paraview"
view <- create_view(paste0("cancer.paraview.", l),
            view.df,
            paste0("c.para.", l))

misty.views <- create_initial_view(intra.df) %>%
  add_views(view)
```

Check whether we can run MISTy.

```{r include=FALSE, eval=FALSE}
x <- run_misty(misty.views) %>%
  collect_results()
plot_improvement_stats(x)
```

Now the same for every sample. Plus views will be saved.

```{r}
if ("cancer.immune.views.RDS" %in% list.files("data")) {
  cancer.immune.views <- readRDS("data/cancer.immune.views.RDS")
} else {
  
  # Set parameters
  l <- 120
  zoi <- 40
  family <- "gaussian"

  # Map over all samples and compute the cancer -> immune view
  cancer.immune.views <- furrr::future_map(unique(meta$SampleID), function(id) {
    
  # Get expression, positions, and types for a sample
  expr <- all.expr[meta$SampleID == id, ] %>% select(11:47)
  pos <- meta %>% filter(SampleID == id) %>% select(row, col)
  types <- meta %>% filter(SampleID == id) %>% pull(Group)
  
  # Get the ids of immune cells
  immune.ids <- which(types == "Immune")
  
  # Get the ids of cancer cells
  cancer.ids <- which(types %in% c("Keratin-positive tumor", "Tumor"))
  
  # Get the distances between all cells
  dists <- distances::distances(as.data.frame(pos))
  
  view.df <- map_dfr(immune.ids, function(i) {
    weights <- get_weight(family, dists[i, ], l, zoi)
    weighted.sum <- colSums(expr[cancer.ids, ] * weights[cancer.ids])
    data.frame(t(weighted.sum))
  })
  
  # Sanity Check for view.df
  # 1. Check for zero variance (otherwise MISTy throws an error)
  target.var <- apply(view.df, 2, stats::sd, na.rm = TRUE)
  view.df <- view.df %>% select(-names(which(target.var == 0)))
  
  # 2. Check for how many unique values
  target.unique <- colnames(view.df) %>%
    purrr::set_names() %>%
    purrr::map_int(~ length(unique(view.df %>% pull(.x))))
  
  view.df <- view.df %>% select(
    names(target.unique[target.unique > cv.folds])
    )
  
  # Sanity Check for intra.df
  # 1. Check for zero variance (otherwise MISTy throws an error)
  intra.df <- expr[immune.ids,]
  target.var <- apply(intra.df, 2, stats::sd, na.rm = TRUE)
  intra.df <- intra.df %>% select(-names(which(target.var == 0)))
  
  # 2. Check for how many unique values
  target.unique <- colnames(intra.df) %>%
    purrr::set_names() %>%
    purrr::map_int(~ length(unique(intra.df %>% pull(.x))))
  
  intra.df <- intra.df %>% select(
    names(target.unique[target.unique > cv.folds])
    )
  
  # Make correct colnames
  colnames(intra.df) <- make.names(colnames(intra.df))
  colnames(view.df) <- make.names(colnames(view.df))
  
  # Create custom "paraview"
  view <- create_view(paste0("cancer.paraview.", l),
              view.df,
              paste0("c.para.", l))
  
  # Add all in return.views
  return.views <- create_initial_view(intra.df) %>%
    add_views(view)

  return.views
  })
  names(cancer.immune.views) <- names(expr.smp)
  saveRDS(cancer.immune.views, "data/cancer.immune.views.RDS")
}
```

## Immune -> Immune

```{r}
if ("immune.immune.views.RDS" %in% list.files("data")) {
  immune.immune.views <- readRDS("data/immune.immune.views.RDS")
} else {
  
  # Set parameters
  l <- 120
  zoi <- 40
  family <- "gaussian"

  # Map over all samples and compute the cancer -> immune view
  immune.immune.views <- furrr::future_map(unique(meta$SampleID), function(id) {
    
  # Get expression, positions, and types for a sample
  expr <- all.expr[meta$SampleID == id, ] %>% select(11:47)
  pos <- meta %>% filter(SampleID == id) %>% select(row, col)
  types <- meta %>% filter(SampleID == id) %>% pull(Group)
  
  # Get the ids of immune cells
  immune.ids <- which(types == "Immune")
  
  # Get the distances between all cells
  dists <- distances::distances(as.data.frame(pos))
  
  view.df <- map_dfr(immune.ids, function(i) {
    weights <- get_weight(family, dists[i, ], l, zoi)
    weighted.sum <- colSums(expr[immune.ids, ] * weights[immune.ids])
    data.frame(t(weighted.sum))
  })
  
  # Sanity Check for view.df
  # 1. Check for zero variance (otherwise MISTy throws an error)
  target.var <- apply(view.df, 2, stats::sd, na.rm = TRUE)
  view.df <- view.df %>% select(-names(which(target.var == 0)))
  
  # 2. Check for how many unique values
  target.unique <- colnames(view.df) %>%
    purrr::set_names() %>%
    purrr::map_int(~ length(unique(view.df %>% pull(.x))))
  
  view.df <- view.df %>% select(
    names(target.unique[target.unique > cv.folds])
    )
  
  # Sanity Check for intra.df
  # 1. Check for zero variance (otherwise MISTy throws an error)
  intra.df <- expr[immune.ids,]
  target.var <- apply(intra.df, 2, stats::sd, na.rm = TRUE)
  intra.df <- intra.df %>% select(-names(which(target.var == 0)))
  
  # 2. Check for how many unique values
  target.unique <- colnames(intra.df) %>%
    purrr::set_names() %>%
    purrr::map_int(~ length(unique(intra.df %>% pull(.x))))
  
  intra.df <- intra.df %>% select(
    names(target.unique[target.unique > cv.folds])
    )
  
  # Make correct colnames
  colnames(intra.df) <- make.names(colnames(intra.df))
  colnames(view.df) <- make.names(colnames(view.df))
  
  # Create custom "paraview"
  view <- create_view(paste0("immune.paraview.", l),
              view.df,
              paste0("i.para.", l))
  
  # Add all in return.views
  return.views <- create_initial_view(intra.df) %>%
    add_views(view)

  return.views
  })
  names(immune.immune.views) <- names(expr.smp)
  saveRDS(immune.immune.views, "data/immune.immune.views.RDS")
}
```

# Running MISTy

## Tumor -> Immune

```{r}
if ("cancer.immune.results.RDS" %in% list.files("data")) {
  cancer.immune.results <- readRDS("data/cancer.immune.results.RDS")
} else {
  cv.folds = 10
  cancer.immune.folders <- map2(
    cancer.immune.views, names(cancer.immune.views), function(smp, name) {
      
      smp %>% run_misty(
        results.folder = paste0("analysis/cell.specific.results/cancer.immune_",
                                name),
                        cv.folds = cv.folds)
    })
  cancer.immune.results <- collect_results(cancer.immune.folders)
  saveRDS(cancer.immune.results, "data/cancer.immune.results.RDS")
}
```

## Immune -> Immune

```{r}
if ("immune.immune.results.RDS" %in% list.files("data")) {
  immune.immune.results <- readRDS("data/immune.immune.results.RDS")
} else {
  cv.folds = 10
  immune.immune.folders <- map2(
    immune.immune.views, names(immune.immune.views), function(smp, name) {
      
      smp %>% run_misty(
        results.folder = paste0("analysis/cell.specific.results/immune.immune_",
                                name),
                        cv.folds = cv.folds)
    })
  immune.immune.results <- collect_results(immune.immune.folders)
  saveRDS(immune.immune.results, "data/immune.immune.results.RDS")
}
```

# MISTy Results

## Tumor -> Immune

### Gain.R2

We see quite some gain in variance explained for IDO which is an immunoregulatory
protein. The same is true for PD-L1

```{r fig.width=6, fig.height=6}
mistyR::plot_improvement_stats(cancer.immune.results, measure = "gain.R2")
```

However, the mean p-values are rather high because there is such a large
variability between the samples.

```{r}
cancer.immune.results$improvements %>% 
  group_by(target, measure) %>%
  summarise(mean = round(mean(value), 2), median = round(median(value), 2),
            .groups = "keep") %>% 
  pivot_longer(cols = c(mean, median), names_to = "summary") %>%
  pivot_wider(names_from = measure, values_from = value) %>%
  select(c(target, summary, gain.R2, gain.RMSE, p.R2, p.RMSE)) %>%
  filter(target %in% c("CD138", "CD209", "CD11c", "CD20", "IDO", "PD.L1", "PD1")) %>%
  knitr::kable()
```

So what happens if we actually stratify the samples here based on the 
classification (e.g. hoping to find a much stronger PD-L1 signal in the
compartimentalized samples).

And in fact it is true that the gain in variance is higher for those
immunoregulatory proteins in compartimentalized tumors.

```{r fig.height=4, fig.width=8}
cancer.immune.results$improvements %>%
  mutate(SampleID = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by = "SampleID") %>%
  filter(measure == "gain.R2") %>%
  filter(Classification != "cold") %>%
  ggplot() +
  geom_boxplot(aes(x = reorder(target, -value), y = value, col = Classification)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Checking the values for immunoregulatory proteins.

First by summarizing using the mean.

```{r}
cancer.immune.results$improvements %>%
  mutate(SampleID = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by = "SampleID") %>%
  filter(Classification != "cold") %>%
  filter(target %in% c("IDO", "PD.L1", "PD1")) %>%
  group_by(target, Classification, measure) %>%
  summarise(mean.measure = round(mean(value, na.rm=TRUE),2),
            .groups="keep") %>%
  pivot_wider(names_from = measure, values_from = mean.measure)
```

And then the median.

```{r}
cancer.immune.results$improvements %>%
  mutate(SampleID = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by = "SampleID") %>%
  filter(Classification != "cold") %>%
  filter(target %in% c("IDO", "PD.L1", "PD1")) %>%
  group_by(target, Classification, measure) %>%
  summarise(mean.measure = round(median(value, na.rm=TRUE), 2),
            .groups="keep") %>%
  pivot_wider(names_from = measure, values_from = mean.measure)
```

Also interestingly the gain in variance explained is negatively correlated
with the mixing score, meaning the more contact there is between immune and
tumor cells the less variance can be explained for IDO.

```{r}
cor(x = cancer.immune.results$improvements %>%
    mutate(SampleID = str_extract(sample, "(?<=_s)[0-9]+")) %>%
    left_join(meta.smp, by = "SampleID") %>%
    filter(Classification != "cold") %>%
    filter(target == "IDO") %>%
    filter(measure == "gain.R2") %>%
    filter(ratio.tumor.immune < 300) %>%
    pull(value),
    y = cancer.immune.results$improvements %>%
        mutate(SampleID = str_extract(sample, "(?<=_s)[0-9]+")) %>%
        left_join(meta.smp, by = "SampleID") %>%
        filter(Classification != "cold") %>%
        filter(target == "IDO") %>%
        filter(measure == "gain.R2") %>%
        filter(ratio.tumor.immune < 300) %>%
        pull(ratio.tumor.immune))

cancer.immune.results$improvements %>%
  mutate(SampleID = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by = "SampleID") %>%
  filter(Classification != "cold") %>%
  filter(target == "IDO") %>%
  filter(measure == "gain.R2") %>%
  filter(ratio.tumor.immune < 300) %>%
  ggplot() +
  geom_point(aes(x=ratio.tumor.immune, y=value)) +
  geom_smooth(aes(x=ratio.tumor.immune, y=value), method="lm", formula = y ~ x)
```

What about the other targets? There are positive and negative correlations
of the gain in variance explained and the tumor-immune mixing score.

```{r}
cancer.immune.results$improvements %>%
  mutate(SampleID = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by = "SampleID") %>%
  filter(measure == "gain.R2") %>%
  filter(ratio.tumor.immune < 300) %>%
  group_by(target) %>%
  summarise(c = cor(value, ratio.tumor.immune), mean.R2.gain = mean(value)) %>%
  filter(mean.R2.gain >= 1) %>%
  ggplot() +
  geom_bar(aes(x=reorder(target, -c), y=c, fill=mean.R2.gain), stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y = "correlation between gain.R2 and mixing score", x = "targets") +
  scale_fill_viridis_c()
```

### View Contributions

```{r fig.width=6, fig.height=6}
mistyR::plot_view_contributions(cancer.immune.results)
```

### Importances

Intraview:

```{r fig.width=8, fig.height=8}
mistyR::plot_interaction_heatmap(cancer.immune.results, view = "intra", 
                                 clean = TRUE)
```

Paraview:

```{r fig.width=8, fig.height=8}
mistyR::plot_interaction_heatmap(cancer.immune.results, view = "c.para.120",
                                 clean = TRUE, cutoff = .8)
```

### Signatures

Performance

```{r}
cancer.immune.perf.sig <- mistyR::extract_signature(cancer.immune.results,
                                                    "performance")
cancer.immune.perf.pca <- stats::prcomp(cancer.immune.perf.sig %>%
                                          select(-sample))
```

```{r}
ggplot() +
  geom_point(aes(x=cancer.immune.perf.pca$x[,1], 
                 y=cancer.immune.perf.pca$x[,2],
                 col=factor(meta.smp$GRADE)))
```

```{r}
fviz_pca_var(cancer.immune.perf.pca, col.var="contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

Importances:

```{r}
cancer.immune.imp.sig <- mistyR::extract_signature(cancer.immune.results,
                                                    "importance")
cancer.immune.imp.pca <- stats::prcomp(cancer.immune.imp.sig %>%
                                          select(-sample))
```

```{r}
ggplot() +
  geom_point(aes(x=cancer.immune.imp.pca$x[,1], 
                 y=cancer.immune.imp.pca$x[,2],
                 col=factor(meta.smp$GRADE)))
```

```{r}
fviz_pca_var(cancer.immune.imp.pca, col.var="contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

## Immune -> Immune

### Gain.R2

```{r fig.width=6, fig.height=6}
mistyR::plot_improvement_stats(immune.immune.results, measure = "gain.R2")
```

### View Contributions

```{r fig.width=6, fig.height=6}
mistyR::plot_view_contributions(immune.immune.results)
```

### Importances

Intraview:

```{r fig.width=8, fig.height=8}
mistyR::plot_interaction_heatmap(immune.immune.results, view = "intra", 
                                 clean = TRUE)
```

Paraview:

```{r fig.width=8, fig.height=8}
mistyR::plot_interaction_heatmap(immune.immune.results, view = "i.para.120",
                                 clean = TRUE, cutoff = .8)
```

### Signatures

Performance

```{r}
immune.immune.perf.sig <- mistyR::extract_signature(immune.immune.results,
                                                    "performance")
immune.immune.perf.pca <- stats::prcomp(immune.immune.perf.sig %>%
                                          select(-sample))
```

```{r}
ggplot() +
  geom_point(aes(x=immune.immune.perf.pca$x[,1], 
                 y=immune.immune.perf.pca$x[,2],
                 col=factor(meta.smp$GRADE)))
```

```{r}
fviz_pca_var(immune.immune.perf.pca, col.var="contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

Importances:

```{r}
immune.immune.imp.sig <- mistyR::extract_signature(immune.immune.results,
                                                    "importance")
immune.immune.imp.pca <- stats::prcomp(immune.immune.imp.sig %>%
                                          select(-sample))
```

```{r}
ggplot() +
  geom_point(aes(x=immune.immune.imp.pca$x[,1], 
                 y=immune.immune.imp.pca$x[,2],
                 col=factor(meta.smp$GRADE)))
```

```{r}
fviz_pca_var(immune.immune.imp.pca, col.var="contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```
