---
title: "Model Comparison"
---

# Set up

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Saez/mibi_analysis/")
```

Loaded packages.

```{r}
suppressPackageStartupMessages(library(mistyR))
suppressPackageStartupMessages(library(future))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(factoextra))
plan("multisession", workers=14)
```

Loading data generated in MIBI_Analysis.Rmd

```{r}
meta <- readRDS("data/meta_data.RDS")
all.expr <- readRDS("data/expression.RDS")
meta.smp <- readRDS("data/sample_meta.RDS")
misty.views.smp <- readRDS("data/misty_views_smp.RDS")
```

# Introduction

To extend the functionality of MISTy a new API was implemented to model
the different views which are ultimatively combined in a linear meta model.
This analysis is supposed to test how these new models work on real-world 
datasets.

# Running MISTy

The MISTy views were generated with the following parameters:

- Intraview: Default
- Juxtaview: l = 40
- Paraview: l = 120, zoi = 40

## Random Forest

```{r}
if ("rf.RDS" %in% list.files("data")) {
  ranger.results <- readRDS("data/rf.RDS")
} else {
  cv.folds = 10
  ranger.results.folders <- map2(
    misty.views.smp, names(misty.views.smp), function(smp, name) {
      
      smp %>% run_misty(results.folder = paste0("analysis/results/ranger_", name),
                        cv.folds = cv.folds)
    })
  ranger.results <- collect_results(ranger.results.folders)
  saveRDS(ranger.results, "data/rf.RDS")
}
```

```{r}
mistyR::plot_improvement_stats(ranger.results)
```

## Bagged MARS (Multivariante Adaptive Regression Splines)

```{r}
if ("bgmars.RDS" %in% list.files("data")) {
  bgmars.results <- readRDS("data/bgmars.RDS")
} else {
  cv.folds = 10
  bgmars.results.folders <- map2(
    misty.views.smp, names(misty.views.smp), function(smp, name) {
      
      smp %>% run_misty(results.folder = paste0("analysis/results/bgmars_", name),
                        model.function = mars_model)
    })
  bgmars.results <- collect_results(bgmars.results.folders)
  saveRDS(bgmars.results, "data/bgmars.RDS")
}
```

```{r}
mistyR::plot_improvement_stats(bgmars.results)
```

## MARS (Multivariante Adaptive Regression Splines)

```{r}
if ("mars.RDS" %in% list.files("data")) {
  mars.results <- readRDS("data/mars.RDS")
} else {
  cv.folds = 10
  mars.results.folders <- map2(
    misty.views.smp, names(misty.views.smp), function(smp, name) {
      
      smp %>% run_misty(results.folder = paste0("analysis/results/mars_", name),
                        model.function = mars_model)
    })
  mars.results <- collect_results(mars.results.folders)
  saveRDS(mars.results, "data/mars.RDS")
}
```

```{r}
mistyR::plot_improvement_stats(mars.results)
```

## Gradient Boosting with Trees

```{r}
if ("tboost.RDS" %in% list.files("data")) {
  tboost.results <- readRDS("data/tboost.RDS")
} else {
  cv.folds = 10
  tboost.results.folders <- map2(
    misty.views.smp, names(misty.views.smp), function(smp, name) {
      
      smp %>% run_misty(results.folder = paste0("analysis/results/tboost_", name),
                        model.function = gradient_boosting_model, booster = "gbtree")
    })
  tboost.results <- collect_results(tboost.results.folders)
  saveRDS(tboost.results, "data/tboost.RDS")
}
```

```{r}
mistyR::plot_improvement_stats(tboost.results)
```

## Gradient Boosting with Linear Models

```{r}
if ("lboost.RDS" %in% list.files("data")) {
  lboost.results <- readRDS("data/lboost.RDS")
} else {
  cv.folds = 10
  lboost.results.folders <- map2(
    misty.views.smp, names(misty.views.smp), function(smp, name) {
      
      smp %>% run_misty(results.folder = paste0("analysis/results/lboost_", name),
                        model.function = gradient_boosting_model, booster = "gblinear")
    })
  lboost.results <- collect_results(lboost.results.folders)
  saveRDS(tboost.results, "data/lboost.RDS")
}
```

```{r}
mistyR::plot_improvement_stats(lboost.results)
```

## Linear Model

```{r}
if ("lm.RDS" %in% list.files("data")) {
  lm.results <- readRDS("data/lm.RDS")
} else {
  cv.folds = 10
  lm.results.folders <- map2(
    misty.views.smp, names(misty.views.smp), function(smp, name) {
      
      smp %>% run_misty(results.folder = paste0("analysis/results/lm_", name),
                        model.function = linear_model)
    })
  lm.results <- collect_results(lm.results.folders)
  saveRDS(lm.results, "data/lm.RDS")
}
```

```{r}
mistyR::plot_improvement_stats(lm.results)
```

# Comparing the results

## Improvements

```{r fig.width=6, fig.height=6}
all.improvements <- map2_dfr(list(ranger.results, bgmars.results, mars.results,
                             tboost.results, lboost.results, lm.results),
                         list("rf", "bgmars", "mars", "tboost", "lboost", "lm"),
                        ~ .x$improvements %>% mutate(algorithm = .y))

all.improvements %>%
  filter(measure == "multi.R2") %>%
  mutate(algorithm = factor(algorithm, levels = c("rf", "tboost", "mars", 
                                                  "bgmars", "lboost", "lm"))) %>%
  group_by(algorithm, target) %>%
  summarise(mean = mean(value), median = median(value), 
            sd = sd(value), .groups = "drop_last") %>%
  ggplot() +
  geom_tile(aes(x = algorithm, y = target, fill = mean)) +
  scale_fill_viridis_c(values = c(0, 0.1, 0.2, 0.4, 0.7, 1)) +
  labs(x = "Algorithm", fill = "Mean Multi.R2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## View Contributions

Random Forest (ranger):

```{r}
mistyR::plot_view_contributions(ranger.results)
```

Tree Boosting (xgboost):

```{r}
mistyR::plot_view_contributions(tboost.results)
```

Bagged MARS (earth):

```{r}
mistyR::plot_view_contributions(bgmars.results)
```

MARS (earth):

```{r}
mistyR::plot_view_contributions(mars.results)
```

Linear Boosting (xgboost):

```{r}
mistyR::plot_view_contributions(lboost.results)
```

Linear Model (lm):

```{r}
mistyR::plot_view_contributions(lm.results)
```
