---
title: "Presentation"
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = "~/Saez/workflowr_projects/mibi_analysis/")

suppressPackageStartupMessages(library(mistyR))
suppressPackageStartupMessages(library(future))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(factoextra))
plan("multisession", workers=14)

# Loading data generated the other tabs.
meta <- readRDS("data/meta_data.RDS")
all.expr <- readRDS("data/expression.RDS")
meta.smp <- readRDS("data/sample_meta.RDS") %>%
    mutate(Classification = ifelse(Classification == "compartimentalized",
                               "compartmentalized", Classification))
misty.views.smp <- readRDS("data/misty_views_smp.RDS")
ranger.results <- readRDS("data/ranger.results.RDS")
cancer.immune.results <- readRDS("data/cancer.immune.RDS")
immune.immune.results <- readRDS("data/immune.immune.RDS")
lmbag.results <- readRDS("data/lmbag.results.RDS")
```

# What is MISTy?

- MISTy is an explainable machine learning framework to probe intra- and 
intercellular marker interactions. 

- Can we predict the expression of marker X using the information about the 
expression of marker Y within a given cell and in the neighboring cells? 

- The assumption being that if this is the case, there is some sort of relationship
between marker X and Y (which has neither to be causal nor linear).

- Highly flexible framework, which allows us for example to specifically look
at the interaction of certain cell types (e.g. How much can the expression of markers
in adjacent tumor cells explain the expression of markers in immune cells).

```{r, echo=FALSE}
# Define variable containing url
url <- "https://www.dropbox.com/s/5wyh520i1wl62ul/graphical_abstract.png?raw=1"
```

<center><img src="`r url`"></center>

# Basic MISTy Workflow

- For each target, first the variance explained (R2) is computed using only 
intracellular information (intraview, here brown) and then information from the spatial 
context (multiview, here green). 

```{r fig.height=4, fig.width=10}
ranger.results$improvements %>%
  filter(measure %in% c("intra.R2", "multi.R2")) %>%
  mutate(measure = factor(measure, levels = c("multi.R2", "intra.R2"))) %>%
  ggplot() +
  geom_boxplot(aes(x=reorder(target, -value), y=value, col=measure),
               width=.6, outlier.shape = NA) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  labs(y = "Statistic", x = "Target") +
  scale_color_manual(values = c("multi.R2" = "forestgreen", 
                                "intra.R2" = "brown"))
```

- If the R2 values for the multiview (gain.R2) is higher, this shows that taking 
the spatial context (expression of all other markers in the neighboring cells) 
has predictive power for the target.

- Thus, we will look at the gain in explained variance across all targets
and samples (see errorbars).

```{r}
plot_improvement_stats(ranger.results, "gain.R2")
```

- We might also be interested in whether the direct neighbors (juxtaview) 
or the broad spatial context (paraview) are more important?

- And most often we see that the broader spatial context (paraview) is more important
than the local cellular niche (juxtaview).

```{r}
plot_view_contributions(ranger.results)
```

- Furthermore, we could ask why there is a large variability
between the samples in the gain in R2 for the different target. What drives
those differences?

- To answer this question we can look at the performance signature, which
is a collection of all performance measures for each target.

```{r}
ranger.perf.sig <- mistyR::extract_signature(ranger.results, "performance")
ranger.perf.sig %>% 
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  mutate_if(is.numeric, round, digits=2)
```

- Doing a PCA on this performance signature clearly shows that MISTy performance
is driven by the tumor structure.

```{r fig.height=5, fig.width=14}
ranger.perf.pca <- stats::prcomp(ranger.perf.sig %>% select(-sample))

p1 <- ggplot() +
  geom_point(aes(x=ranger.perf.pca$x[,1], y=ranger.perf.pca$x[,2],
                 col=factor(meta.smp$Classification))) +
  labs(col = "Classification", x="PCA-1", y="PCA-2")

p2 <- ggplot() +
  geom_point(aes(x=ranger.perf.pca$x[,1], 
                 y=ranger.perf.pca$x[,2],
                 col=log10(meta.smp$ratio.tumor.immune))) +
  scale_color_viridis_c() +
  labs(col = "Log10 Mixing Score", x="PCA-1", y="PCA-2")

gridExtra::grid.arrange(p1, p2, ncol=2)
```

- Alternatively one could look at the boxplot containing all gain R2 values.

- Especially for cold tumors, the gains in R2 are rather low and according
to the plot the overall differences between compartmentalized and
mixed tumors are not too big.

```{r fig.width=6, fig.height=3}
ranger.results$improvements %>%
  filter(measure == "gain.R2") %>% 
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  ggplot() +
  geom_boxplot(aes(x=Classification, y=value, col=Classification)) +
  scale_color_manual(values = c("cold" = "red",
                                "compartmentalized" = "forestgreen", 
                                "mixed" = "blue")) +
  labs(y = "Gain R2")
```

- So for which targets is the gain in R2 different between
compartmentalized and mixed samples? Meaning for the prediction 
of which targets is it important whether the special context comprises
a structured or unstructured TME?

- On top we can see the targets for which the gain in R2 is significantly 
different between compartmentalized and mixed samples. On the bottom the
multi R2 (total explained variance) is shown.

```{r fig.height=3, fig.width=7}
tmp <- ranger.results$improvements %>%
  filter(measure == "gain.R2") %>% 
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by=c("sample" = "SampleID"))

targets <- tmp$target %>% unique

# t.test for all targets
all.targets <- map(targets, function(t) {
  tmp <- ranger.results$improvements %>%
    filter(measure == "gain.R2") %>% 
    mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
    left_join(meta.smp, by=c("sample" = "SampleID")) %>%
    filter(target == t)
  
  test.results <- t.test(x = (tmp %>% filter(Classification=="compartmentalized") %>%
              pull(value)),
       y = (tmp %>% filter(Classification=="mixed") %>%
              pull(value)),
       alternative = "two.sided")
  
  list("target" = t, "t.test" = test.results)
})

# looking at the boxplots for each target
sign.targets <- map(all.targets, function(target.test) {
  if (target.test$t.test$p.value < 0.05) {
    return(target.test$target)
  }
}) %>% unlist

ranger.results$improvements %>%
  filter(measure == "gain.R2") %>% 
  mutate(measure = ifelse(measure == "gain.R2",
                          "Gain Explained Variance (R2) [%]",
                          "Total Explained Variance (R2)")) %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  filter(target %in% sign.targets) %>%
  filter(Classification != "cold") %>%
  mutate(Classification = factor(Classification)) %>%
  ggplot() +
  geom_boxplot(aes(x=target, y=value, col=Classification)) +
  scale_color_manual(values = c("compartmentalized" = "forestgreen", "mixed" = "blue"))
```

- But what kind of interactions could explain those differences? Which other
markers are important for the prediction of those targets?

- Let's look at the **importances** of the **paraview** for these targets in 
compartmentalized and mixed samples.

```{r fig.width=5, fig.height=5}
cutoff = 0.5
df <- ranger.results$importances %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  filter(Target %in% sign.targets) %>%
  filter(Classification != "cold") %>%
  group_by(Classification, view, Predictor, Target) %>%
  summarise(m.imp = mean(Importance), .groups="drop") %>%
  filter(view == "para.120") %>%
  mutate(Importance = ifelse(m.imp < cutoff, 0, m.imp))

clean.predictors <- df %>%
  group_by(Predictor) %>%
  summarise(test = sum(Importance)) %>%
  filter(test > 1e-3) %>%
  pull(Predictor) 

df %>%
  filter(Predictor %in% clean.predictors) %>%
  ggplot(., aes(x = Predictor,y = Target)) +
  geom_tile(aes(fill = Importance), color = "black") +
  scale_fill_gradient2(low = "white", mid = "white", high = "#8DA0CB", midpoint = cutoff) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  coord_equal() +
  facet_wrap(~ Classification, ncol=1)
```

- Now we could also condense this down and only look at the contrast. 

```{r fig.width=5, fig.height=5}
cutoff <- 0.5
ranger.results$importances %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  filter(Target %in% sign.targets) %>%
  filter(Classification != "cold") %>%
  group_by(Classification, view, Predictor, Target) %>%
  summarise(Importance = mean(Importance), .groups="drop") %>%
  filter(view == "para.120") %>%
  filter(Predictor %in% clean.predictors) %>%
  pivot_wider(names_from=Classification, values_from=Importance) %>%
  group_by(Predictor, Target) %>%
  summarise(c = sum(compartmentalized, na.rm = TRUE), m = sum(mixed, na.rm = TRUE),
            .groups="drop") %>%
  mutate(Contrast.Compartmentalized.Mixed = c - m,
         Contrast.Mixed.Compartmentalized = m -c) %>%
  mutate(Contrast.Compartmentalized.Mixed = 
           ifelse(Contrast.Compartmentalized.Mixed < cutoff, 0,
                  Contrast.Compartmentalized.Mixed)) %>%
  mutate(Contrast.Mixed.Compartmentalized = 
           ifelse(Contrast.Mixed.Compartmentalized < cutoff, 0,
                  Contrast.Mixed.Compartmentalized)) %>%
  pivot_longer(cols = starts_with("Contrast"), values_to="Importance") %>%
  mutate(name = ifelse(name == "Contrast.Compartmentalized.Mixed",
                       "Contrast: Compartmenalized - Mixed",
                       "Contrast: Mixed - Compartimenalized")) %>%
  ggplot(., aes(x = Predictor,y = Target)) +
  geom_tile(aes(fill = Importance), color = "black") +
  scale_fill_gradient2(low = "white", mid = "white", high = "#8DA0CB", midpoint = cutoff) +
  facet_wrap(~ name, ncol=1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  coord_equal()
```

- For example EGFR and FoxP3 are more important for prediction of these
targets in mixed samples, whereas CD45 is more important in compartmentalized
samples.

# Neighborhood Composition Analysis

- Since MISTy is a flexible framework, I want to give another short example
how we can use MISTy to predict the identity of cells using the identity
of the 50 nearest-neighbor of a cell.

- Here I excluded Tumor cells and cells which are not really present in 
every sample.

- Let's start again by looking at the gain in R2.

```{r}
notumor.misty.results <- readRDS("data/no.tumor.compositon.RDS")
plot_improvement_stats(notumor.misty.results, "gain.R2")
```

- As expecpted these results are driven by the tumor structure:

```{r}
notumor.misty.results$improvements %>%
  mutate(sample = str_extract(sample, "(?<=smp.)[0-9]+")) %>%
  filter(measure == "gain.R2") %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  ggplot() +
  geom_boxplot(aes(x=target, y=value, col=Classification)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_color_manual(values = c("compartmentalized" = "forestgreen", 
                                "mixed" = "blue")) 
```

- But in general what kind of interactions explain those results?

- We can see kind of see there is a structure in the immune response. The levels
of macrophages are informative about the presence/absense of other cell types
(also macrophages are by far the most abundant immune cell type).

- However, interpreting this plot is not straight forward as those interaction
can be very complex (non-linear)

```{r}
plot_interaction_heatmap(notumor.misty.results, "para.50", cutoff = 0)
```

- For example we see the mutual interaction between CD4 T-cells and B-cells, 
which we could investigate by actually looking at the samples (which have
rather high numbers of both cell types).

- It seems like there is some coordinated though not in all samples!

```{r fig.width=12, fig.height=8}
df <- meta %>%
  group_by(SampleID) %>%
  count(immuneGroup) %>%
  ungroup() %>%
  pivot_wider(names_from=immuneGroup, values_from=n)

q75_cd4 <- quantile(df$`CD4 T`, .6, na.rm=TRUE)
q75_b <- quantile(df$B, .6, na.rm=TRUE)

sample.oi <- df %>%
  filter(`CD4 T` >= q75_cd4, B >= q75_b) %>%
  pull(SampleID)

meta %>%
  mutate(SampleID = as.character(SampleID)) %>%
  left_join(meta.smp, by="SampleID") %>%
  filter(SampleID %in% sample.oi) %>%
  mutate(class = case_when(
    Group %in% c("Keratin-positive tumor", "Tumor") ~ "Tumor",
    !(immuneGroup %in% c("B", "CD4 T")) ~ "Ignore",
    TRUE ~ immuneGroup
  )) %>%
  filter(class != "Ignore") %>%
  ggplot() +
  geom_point(aes(x=row, y=col, col=class), size=.1) +
  facet_wrap(~ SampleID, ncol=4) +
  guides(colour = guide_legend(override.aes = list(size = 2)))
```

# Cell Type Specific Analysis

- Another use case might be to probe the interactions between certain cell types.

- For example predicting the expression of markers in immune cells using 
information from the tumor cell in the broader tissue structure (paraview).

## Tumor -> Immune

- We can see again that the expression of the tumor cells are informative.

- We also see that there is quite some gain in variance explained for
IDO, PD_L1 and PD1, so which interactions could explain that?

```{r}
plot_improvement_stats(cancer.immune.results, "gain.R2")
```

- Thus, we will have a look at the corresponding importances.

- There is no single important predictor, rather many predictors are
important for every immuno-regulatory proteins

- But some predictors are specific, such as EGFR for PD-L1.

```{r fig.height=3, fig.width=6}
cutoff = 0
cancer.immune.results$importances %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  filter(Target %in% c("IDO", "PD.L1", "PD1", "Lag3")) %>%
  group_by(Classification, view, Predictor, Target) %>%
  summarise(m.imp = mean(Importance), .groups="drop") %>%
  filter(view == "c.para.120") %>%
  mutate(Importance = ifelse(m.imp < cutoff, 0, m.imp)) %>%
  ggplot(., aes(x = Predictor,y = Target)) +
  geom_tile(aes(fill = Importance), color = "black") +
  scale_fill_gradient2(low = "white", mid = "white", high = "#8DA0CB", midpoint = cutoff) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  coord_equal()
```

- And we could actually ask the same question when looking at the interaction
of immune cells.

- What kind of interactions explain 

## Immune -> Immune



- What kind of interactions are important between immune cells?

- For example PD1 expression is helpful to predict Lag3 expression.

```{r fig.height=3, fig.width=6}
cutoff = 0
immune.immune.results$importances %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  filter(Target %in% c("IDO", "PD.L1", "PD1", "Lag3")) %>%
  group_by(Classification, view, Predictor, Target) %>%
  summarise(m.imp = mean(Importance), .groups="drop") %>%
  filter(view == "i.para.120") %>%
  mutate(Importance = ifelse(m.imp < cutoff, 0, m.imp)) %>%
  ggplot(., aes(x = Predictor,y = Target)) +
  geom_tile(aes(fill = Importance), color = "black") +
  scale_fill_gradient2(low = "white", mid = "white", high = "#8DA0CB", midpoint = cutoff) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  coord_equal()
```

# Thoughts

- We are happy to collaborate

- Since MISTy was rather developed to probe intra- and especially intercellular
interactions it would be helpful to have more markers that help to decipher
pathway activity and expression of ligands.

- How does the site/location in the solid tumor where the sample is taken from influence the sample-to-sample heterogeneity? Did you by now had a look at similar samples, 
and are the effects you saw preserved across different sites in the same
patient?

# Appendix

- Interestingly the prediction of the other immuno-regulatory proteins apart from
PD-1 did not depend on the tumor structure.

```{r}
ranger.results$improvements %>%
  filter(measure == "gain.R2") %>% 
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  filter(target %in% c("IDO", "PD.L1", "PD1", "Lag3")) %>%
  filter(Classification != "cold") %>%

  ggplot() +
  geom_boxplot(aes(x=target, y=value, col=Classification)) +
  scale_color_manual(values = c("compartmentalized" = "forestgreen", "mixed" = "blue")) +
  labs(x = "Target", y = "Gain Explained Variance (R2)")
```

- One could also run MISTy with linear models as base learner in a bagged to
to increase interpretability. Though this strongly reduced the variance that
can be explained.

```{r fig.height=4, fig.width=10}
df1 <- ranger.results$improvements %>%
  filter(measure == "gain.R2") %>%
  mutate(method = "ranger")

df2 <- lmbag.results$improvements %>%
  filter(measure == "gain.R2") %>%
  mutate(method = "lm")

rbind(df1, df2) %>%
  mutate(method = factor(method, levels = c("ranger", "lm"))) %>%
  ggplot() +
  geom_boxplot(aes(x=reorder(target, -value), y=value, col=method),
               outlier.shape = NA, width = .5) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  labs(y = "Statistic", x = "Target") +
  scale_color_manual(values = c("ranger" = "forestgreen", 
                                "lm" = "brown"))
```

- The interactions we have looked at before.

```{r fig.width=7, fig.height=7}
cutoff = 0.5
lmbag.results$importances %>%
  mutate(sample = str_extract(sample, "(?<=_s)[0-9]+")) %>%
  left_join(meta.smp, by=c("sample" = "SampleID")) %>%
  filter(Target %in% sign.targets) %>%
  filter(Classification != "cold") %>%
  group_by(Classification, view, Predictor, Target) %>%
  summarise(m.imp = mean(Importance), .groups="drop") %>%
  filter(view == "para.120") %>%
  mutate(Importance = ifelse(m.imp < cutoff, 0, m.imp)) %>%
  ggplot(., aes(x = Predictor,y = Target)) +
  geom_tile(aes(fill = Importance), color = "black") +
  scale_fill_gradient2(low = "white", mid = "white", high = "#8DA0CB", midpoint = cutoff) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  coord_equal() +
  facet_wrap(~ Classification, ncol=1)
```

- The grades are not really associated with the performance signature of MISTy 
in contrast to the tumor structure as seen before.

```{r fig.width=8, fig.height=6}
p1 <- ggplot() +
  geom_point(aes(x=ranger.perf.pca$x[,1], y=ranger.perf.pca$x[,2],
                 col=factor(meta.smp$GRADE))) +
  labs(col = "Grade", x="PCA-1", y="PCA-2")

p2 <- ggplot() +
  geom_point(aes(x=ranger.perf.pca$x[,1], y=ranger.perf.pca$x[,2],
                 col=factor(meta.smp$STAGE))) +
  labs(col = "Stage", x="PCA-1", y="PCA-2")

p3 <- ggplot() +
  geom_point(aes(x=ranger.perf.pca$x[,1], y=ranger.perf.pca$x[,2],
                 col=factor(meta.smp %>%
                              mutate(STAGE = substring(STAGE, 1, 1)) %>%
                              pull(STAGE)))) +
  labs(col = "red. Stage", x="PCA-1", y="PCA-2")

p4 <- ggplot() +
  geom_point(aes(x=ranger.perf.pca$x[,1], y=ranger.perf.pca$x[,2],
                 col=(meta.smp %>%
  mutate(TIL_score = ifelse(TIL_score %in% c("NA", "NaN"), 
                            NA, TIL_score)) %>% pull(TIL_score) %>%
    as.numeric()))) +
  labs(col = "TIL Score", x="PCA-1", y="PCA-2") +
  scale_color_viridis_c()
gridExtra::grid.arrange(p1, p2, p3, p4, ncol=2)
```

- But what about the importance signatures? I am not quite sure whether there
are patterns, once could argue about the grade I guess.

```{r fig.width=8, fig.height=6}
ranger.imp.sig <- mistyR::extract_signature(ranger.results, "importance")
ranger.imp.pca <- stats::prcomp(ranger.imp.sig %>% select(-sample))

p1 <- ggplot() +
  geom_point(aes(x=ranger.imp.pca$x[,1], y=ranger.imp.pca$x[,2],
                 col=factor(meta.smp$GRADE))) +
  labs(col = "Grade", x="PCA-1", y="PCA-2")

p2 <- ggplot() +
  geom_point(aes(x=ranger.imp.pca$x[,1], y=ranger.imp.pca$x[,2],
                 col=factor(meta.smp$STAGE))) +
  labs(col = "Stage", x="PCA-1", y="PCA-2")

p3 <- ggplot() +
  geom_point(aes(x=ranger.imp.pca$x[,1], y=ranger.imp.pca$x[,2],
                 col=factor(meta.smp %>%
                              mutate(STAGE = substring(STAGE, 1, 1)) %>%
                              pull(STAGE)))) +
  labs(col = "red. Stage", x="PCA-1", y="PCA-2")

p4 <- ggplot() +
  geom_point(aes(x=ranger.imp.pca$x[,1], y=ranger.imp.pca$x[,2],
                 col=meta.smp %>%
  mutate(TIL_score = ifelse(TIL_score %in% c("NA", "NaN"), 
                            NA, TIL_score)) %>% pull(TIL_score) %>%
    as.numeric())) +
  labs(col = "TIL Score", x="PCA-1", y="PCA-2") +
  scale_color_viridis_c()
gridExtra::grid.arrange(p1, p2, p3, p4, ncol=2)
```




